---
title: "Bayesian exposure-response modeling for binary data"
output:
  slidy_presentation:
#    theme: cerulean
    fig_width: 9
    fig_height: 5
    font_adjustment: 3
    transition: none
    css: slidystyles.css
    footer: Metrum Research Group &copy 2021
    mathjax: local
    self_contained: false
    df_print: tibble
#bibliography: alpha.bib
#nocite: | 
#    @2795, @2192, @3684
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
library(tidyverse)
library(stringr)
library(haven)
library(survival)
library(GGally)
library(binom)
library(texreg)
library(brms)
library(tidybayes)
library(DHARMa)
library(mgcv)

set.seed(314159)

theme_set(theme_bw())

expit <- function(x) 1 / (1+exp(-x))
``` 

# Outline

* Brief introduction to Bayesian analysis
* Fitting models
* Model checking


# Approach to Bayesian modeling in this course

* For this series of classes we are going to use Stan to do Bayesian modeling

    * Stan is a probabilistic programming language for fitting Bayesian models.  
    * By default it uses Hamiltonian Monte Carlo (HMC), specifically a variation called the no U-turn sampler (NUTS).
    * We will go into more details about HMC/NUTS later in the course

* Start with using `brms` as our gateway to Stan 
    * brms is a package that enables simple fitting of many types models that you can fit using `glm`, `survreg`, `lme`, `nlme`, etc.
    * Allows quick access to Bayesian inference

* Later we will program our own Stan models and run them with `cmdstanr`
    * Learn more about the Stan language
    * Fit models that are not supported in `brms`


# A brief review of Bayesian inference


$\color{firebrick}{\text{Bayes Rule}}$ the basis for inference about model parameters ($\theta$) given data ($y$) and prior knowledge about model parameters ($p\left(\theta\right)$):
\begin{eqnarray*}
p\left(\theta\,|\,y\right) &=& \frac{p\left(\theta\right)p\left(y\,|\,\theta\right)}{p\left(y\right)}
= \frac{p\left(\theta\right)p\left(y\,|\,\,\theta\right)}{\int{p\left(\theta\right)p\left(y\,|\,\theta\right) d\theta}}
\\
&\propto& p\left(\theta\right)p\left(y\,|\,\theta\right)
\end{eqnarray*}
<!-- The $p$'s are probabilities or probability densities of the specified random variables. -->

:::{.notebox}
**Goals: **

* Inference about $\theta$ or a function of $\theta$
* Predictions of future observations
:::

* The posterior summarizes what we know about $\theta$, but typically we can't express $p\left(\theta\,|\,y\right)$ in closed-form.

    * We'll use Markov Chain Monte Carlo to obtain samples from $p\left(\theta\,|\,y\right)$.


# Bayesian modeling/inference process using MCMC

<font size="6">

1. Construct a model for the data, conditional on parameters $\theta$, $p\left(y \,|\, \theta\right)$

2. Construct a prior distribution for $\theta$, $p\left(\theta\right)$

    * Ideally based on all available evidence/knowledge (or belief)
    * Or deliberately select a non-informative (or weakly informative)
  prior

3.  Sample from the posterior distribution for $\theta$, $p\left(\theta \,|\, y\right)$.

    * Look at convergence and sampler diagnostics
    * Use for inferences regarding parameter values

4. Sample from the posterior predictive distribution for $y_\text{new}$  $p\left(y_\text{new} \,|\, y\right) = \int{p\left(y_{new}\,|\,\theta\right)p\left(\theta\,|\,y\right) d\theta}$.

    * Use for inferences regarding future observations

<!-- $$p\left(y_{new}\,|\,y\right) = \int{p\left(y_{new}\,|\,\theta\right)p\left(\theta\,|\,y\right) d\theta}$$ -->
</font>

# Bayesian ingredients for MCMC sampling from a posterior

* Data
* Model for the outcome(s) -- the likelihood
* Models for the parameters -- the prior distribution
* MCMC tool -- Stan (via `brms` or `cmdstanr`)

# Ingredients for HMC/NUTS

- A starting point in the parameter space (initial value, one per chain)
- Number of MCMC samples used to tune the HMC/NUTS algorithm (`warmup`)
    * This is not exactly the same as the burn-in in other MCMC algorithms
    * NUTS uses these samples to [adaptively tune the sampler](https://mc-stan.org/docs/2_26/reference-manual/hmc-algorithm-parameters.html)
    
- Total number of samples to take, including the warm-up (`iter`)
- Parameters which inform how the sampler should adapt
    * Defaults are usually good for 'simple' models
    * Often need to modify for more complex, hierarchical models
    * Return to these later in the course


# Let's re-fit our AE model using `brms`

```{r,echo=FALSE}
load('../data/aedat.RDS')
source('preprocess_data.R')
```


```{r, eval=FALSE}
mod01_glm <- glm(formula = AE01 ~ CAVGSS + BWT + PTTYPE + SEXTXT,  
                  data = aedat, 
                  family = bernoulli(link = "logit")
                 )
```

<br>

```{r, eval=FALSE}
mod01_stan <- brm(formula = AE01 ~ CAVGSS + BWT + PTTYPE + SEXTXT,  
                  data = aedat, 
                  family = bernoulli(link = "logit"),
                  warmup = 500, 
                  iter = 2000, 
                  chains = 4, 
                  inits = "random", 
                  cores = 4,
                  seed = 123)
```

# What about the prior distributions?

* By default, `brms` uses flat, non-informative prior distributions for regression coefficients
* We can specify priors directly through the `prior` argument.
  * More to come in a few slides

# Model summary

```{r, echo=FALSE, eval=TRUE}
mod01_stan <- brm(formula = AE01 ~ CAVGSS + BWT + PTTYPE + SEXTXT,  
                  data = aedat, 
                  family = bernoulli(link = "logit"),
                  warmup = 500, 
                  iter = 2000, 
                  chains = 4, 
                  inits = "random", 
                  cores = 4,
                  seed = 123,
                  file = 'mod01_stan_output')
```

```{r}
summary(mod01_stan)
```

* Defaults: Estimate = posterior mean and Est.Err = posterior sd.  Can modify these using the `robust` argument to `brm`

# MCMC convergence diagnostics

* Traceplots
    * Plot of sampled values vs iteration 
    * Look for stationarity and good mixing: fuzzy caterpillar

* $\hat{R}$
    * Heuristically: $\frac{\text{total variance of } \theta \text{ (between and within-chain)}}{\text{average within-chain variance of } \theta}$
    * Target: $\hat{R} < 1.05$
    * Output: Summary and plot (`mcmc_plot(mod01_stan, type='rhat')`)

* Effective sample sizes
    * bulk ESS for assessing posterior means, medians, etc
    * tail ESS for assessing tail percentiles (5th, 95th)
    * Target: Depends on your goals
    * Output: Summary
    
# Traceplots

```{r}
bayesplot::mcmc_trace(mod01_stan)
```

# Let's see the default priors in our model

```{r}
prior_summary(mod01_stan)
```

# BRMS centers all predictors in the model

Mathematically, the RHS of the model `y ~ x1 + x2` is 
$$b\_Intercept + b\_x1 \cdot x1 + b\_x2 \cdot x2$$

Or, equivalently,
$$Intercept + b\_x1 \cdot (x1 - \overline{x1}) + b\_x2 \cdot (x2 - \overline{x2})$$
where 
$$b\_Intercept = Intercept - b_x1 \cdot \overline{x1} - b\_x2 \cdot \overline{x2}$$

This is the parameterization that `brms` uses.  (***Why do you think that is?***)

We need to specify priors for `Intercept`, `b_x1` and `b_x2`


# To change them use the `set_priors` function

A Normal(mean=0, sd=5) prior on all covariate effects:
```{r}
priors_mod01 <- set_prior('normal(0,5)', class='b')
```

<br>

A Normal(0,5) prior on CAVGSS and a DoubleExponential prior on the other effects:
  
```{r}
priors_mod01_de <- set_prior('normal(0,5)', class='b', coef='CAVGSS') +
  set_prior('double_exponential(0,1)', class='b')
```

See [Stan functions reference](https://mc-stan.org/docs/2_26/functions-reference/index.html) for list of all available distributions.

# Workbook Bayes01

* Model fitting in `brms`
* Convergence diagnostics

# Model diagnostics

* We can use similar diagnostics as with likelihood based methods, but now using posterior predictive distributions

    * Quantile residual plots
    * Posterior predictive checks

# Quantile residuals

1. Simulate from posterior predictive distribution
    * `brms` has a **`posterior_predict`** function to generate samples in a matrix
2. Use `createDHARMa` function in the `DHARMa` package to format output
    * Input is a matrix of posterior samples and the observed outcome data
3. Make plots as before
    * Residuals vs predicted
    * Residuals vs predictors

# DHARMa examples

::::{.columns}

:::{.column width=47.5%}
```{r, echo=TRUE, out.width="120%"}
postpred_sample_mod01 = posterior_predict(mod01_stan)

dharma_resids = createDHARMa(simulatedResponse = t(postpred_sample_mod01),
                             observedResponse = aedat$AE01,
                             integerResponse = TRUE)

plot(dharma_resids)
```

:::

:::{.column width=5%}
:::

:::{.column width=47.5%}

```{r, out.width="90%"}
aedat %>% ungroup() %>% 
  mutate(quantile_residual = dharma_resids$scaledResiduals) %>% 
  ggplot(aes(x=BWT, y=quantile_residual)) +
  geom_point() +
  geom_smooth()
```

:::

::::

# Posterior predictive checks

1. Simulate from posterior predictive distribution
    * `tidy_bayes` has an **`add_predicted_draws`** function to append the samples to a data frame
2. Compute some summary statistic on each posterior draw and on the observed data
    * Summary statistic depends on what you want to diagnose
    * For binary models, it is almost always the expected value (mean)
3. Plot distribution of summary statistics and overlay observed values
    * Type of plot depends on grouping factor and summary statistic


# Simulate from posterior predictive distribution


```{r}
aedat_pp = add_predicted_draws(model=mod01_stan, newdata = aedat)

aedat_pp %>% ungroup() %>% 
  select(USUBJID, PTTYPE, AE01, Quartile:.prediction) %>% 
  slice_tail(n=4)
```

# Plot the VPC

::::{.columns}

:::{.column}

```{r, 'bayes-vpc', fig.show = 'hide'}
# Observed data summary
obs_summary <- aedat %>% 
  group_by(Quartile) %>% 
  summarise(phat_obs = mean(AE01))

#Simulated data summary
sim_summary <- aedat_pp %>% 
  group_by(.draw,Quartile) %>% 
  summarise(phat_sim = mean(.prediction))

# VPC
sim_summary %>% 
  ggplot(aes(x=Quartile, y=phat_sim)) +
  geom_violin() +
  geom_point(data=obs_summary, aes(y=phat_obs), col='red',shape=3) +
  labs(x='', y='Proportion with AE') +
  coord_flip()
```
:::

:::{.column}
```{r, ref.label='bayes-vpc', echo=FALSE}

```

:::

::::


# VPC for continuous variable: define summary statistic

* Fit generalized additive model (smoother) to each simulated dataset
* Predict at a fixed grid of values (0 to 95$^{th}$ percentile)

```{r}
summary_function <- function(.data, .x_name, .y_name='value') {
  .data <- .data %>% ungroup() %>% rename('xvar' = .x_name, 'yvar'=.y_name)
  x_grid <- with(.data, seq(from = min(xvar),
                           to = quantile(xvar,probs = 0.95), 
                           length = 100))
  fit <- gam(yvar ~ s(xvar), family=binomial(link='logit'), data=.data)
  predictions <- predict(fit, newdata = data.frame(xvar=x_grid), type='response')
  return( data.frame(xvar=x_grid, prediction = predictions))
}
```

# Compute summary statistics for observed data

```{r}
obs_summary <- summary_function(aedat, .x_name = 'CAVGSS', .y_name='AE01') %>% 
  mutate(type='Observed')

head(obs_summary)
```

# Compute summary on each simulated study

```{r}
sim_summary <- aedat_pp %>% 
  # Nest everying except the simulation name
  nest(cols=-.draw) %>% 
  # Use 200 sims for demonstration
  sample_n(size=200) %>% 
  # Compute summary stats for each simulated dataset
  mutate(predictions = map(cols, ~summary_function(.x, 
                                                   .x_name='CAVGSS',
                                                   .y_name='.prediction'))) %>% 
  select(.draw,predictions) %>% 
  unnest(cols=predictions) %>% 
  # Summarise across simulated data sets
  group_by(xvar) %>% 
  summarise(qlo = quantile(prediction, probs = 0.05),
            qhi = quantile(prediction, probs = 0.95),
            prediction=median(prediction)
            ) %>% 
  mutate(type = 'Simulated')
```

# Plot VPC

```{r, 'cont-vpc'}
sim_summary %>% bind_rows(obs_summary) %>% 
  ggplot(aes(x=xvar, y=prediction)) +
  geom_line(aes(col=type, group=type)) +
  geom_ribbon(aes(ymin=qlo, ymax=qhi, fill=type), alpha=0.2) +
  labs(x='Steady-state Cavg', y='Probability of severe AE')

```

# Model Comparison

* Goal: maximize expected log predictive density (ELPD) for future data
    * This is a measure of <red>out-of-sample</red> prediction quality

* Leave-one-out cross-validation (LOO-CV) to approximate ELPD
    * Involves fitting N models
    * Can be approximated using pareto smoothed importance sampling of the posterior samples
    * `loo` package
    
* WAIC also approximates -2 ELPD
    * Proven to be asymptotocally equivalent to LOO-CV (modulo the $-2$)
    * Lower is better
    * `loo` package

# Workbook Bayes02: Model evaluation and comparison

# References
