---
title: "Workbook"
author: "Binomial Density"
date: "`r Sys.Date()`"
output:
  html_document:
    # css: docs/src/styles/styles.css
    number_sections: true
    theme: united
    toc: true
    toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(haven)
library(GGally)
library(binom)
library(texreg)
library(mgcv)
library(DHARMa)

expit <- function(x) 1 / (1+exp(-x))
```

# Workbook: Exploring the binomial density

## rbinom

```{r, "TRY rbinom", purl=TRUE, results = "hide"}
set.seed(314159)
y <- rbinom(n=20, size=1, prob=0.5)
table(y)
```

Exercise:

* Generate 100 independent Bernoulli observations with success probability = 0.2.
* Do the same thing again, both with and without re-setting the "seed"

```{r}

```


## dbinom

```{r, "dbinom", purl=TRUE, results = "hide"}
set.seed(1)
pp <- 0.7 # probability that event happens
y <- 1 # indicates event occured
nn <- 1 # number of "attempts"
pp^y*(1-pp)^(nn-y)
dbinom(x=y, size=nn, prob=pp) # same thing
y <- c(0, 0, 0, 1, 0, 1)
# Likelihood contributions, evaluated at p = 0.1
dbinom(x=y, size=1, prob=0.1)
```


## Likelihood computation

```{r, "TRY dbinom", purl=TRUE, results = "hide"}
set.seed(1)
# Joint likelihood, evaluated at p = 0.1
prod(dbinom(x=y, size=1, prob=0.1))
```

Exercise do one of the following:

* Use trial and error to find the probability that maximizes the joint likelihood.
* Use some more clever trick to accomplish the same goal.

```{r}

```


## Likelihood maximization

Advanced exercise: look up `?optim` and use that to maximize the joint likelihood

```{r}

```

