---
title: "Workbook"
author: "Binomial Density"
date: "`r Sys.Date()`"
output:
  html_document:
    # css: docs/src/styles/styles.css
    number_sections: true
    theme: united
    toc: true
    toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(haven)
library(GGally)
library(binom)
library(texreg)
library(mgcv)
library(DHARMa)

expit <- function(x) 1 / (1+exp(-x))
```

# Workbook: Exploring the binomial density

## rbinom

```{r, "TRY rbinom", purl=TRUE, results = "hide"}
set.seed(314159)
y <- rbinom(n=20, size=1, prob=0.5)
table(y)
```

Exercise:

* Generate 100 independent Bernoulli observations with success probability = 0.2.
* Do the same thing again, both with and without re-setting the "seed"

```{r}
set.seed(1)
rbinom(n=100, size=1, prob=0.2)
rbinom(n=100, size=1, prob=0.2)
set.seed(1)
rbinom(n=100, size=1, prob=0.2)
```


## dbinom

```{r, "dbinom", purl=TRUE, results = "hide"}
set.seed(1)
pp <- 0.7 # probability that event happens
y <- 1 # indicates event occured
nn <- 1 # number of "attempts"
pp^y*(1-pp)^(nn-y)
dbinom(x=y, size=nn, prob=pp) # same thing
y <- c(0, 0, 0, 1, 0, 1)
# Likelihood contributions, evaluated at p = 0.1
dbinom(x=y, size=1, prob=0.1)
```


## Likelihood computation

```{r, "TRY dbinom", purl=TRUE, results = "hide"}
set.seed(1)
# Joint likelihood, evaluated at p = 0.1
prod(dbinom(x=y, size=1, prob=0.1))
```

Exercise do one of the following:

* Use trial and error to find the probability that maximizes the joint likelihood.
* Use some more clever trick to accomplish the same goal.



```{r}
# Trial and error evaluation of joint likelihood: 
prod(dbinom(x = y, size = 1, prob = 0.1))
prod(dbinom(x = y, size = 1, prob = 0.2))
prod(dbinom(x = y, size = 1, prob = 0.3))
prod(dbinom(x = y, size = 1, prob = 0.4))
# Maximum seems to be somewhere near p = 0.3
```

```{r}
# More clever approach using grid search 
joint_likelihood <- function(p) prod(dbinom(x = y, size = 1, prob = p))
pdat <- tibble(p = seq(0, 1, by = 0.01)) %>% 
  rowwise() %>% 
  mutate(lik = joint_likelihood(p))
pdat$p[which.max(pdat$lik)]
```


## Likelihood maximization

Advanced exercise: look up `?optim` and use that to maximize the joint likelihood

```{r}
optim(p = 0.5, # initial value
      joint_likelihood, # function to be maximized
      method = "Brent", # per help page, use this for 1d problems
      lower = 0, upper = 1, # solution must be between 0 and 1
      control = list(fnscale = -1) # maximize rather than minimize
      )$par
```

