---
title: "Workbook"
author: "Odds ratios, relative risks, and the logit transformation"
date: "`r Sys.Date()`"
output:
  html_document:
    # css: docs/src/styles/styles.css
    number_sections: true
    theme: united
    toc: true
    toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(haven)
library(GGally)
library(binom)
library(texreg)
library(mgcv)
library(DHARMa)

expit <- function(x) 1 / (1+exp(-x))
```




# Workbook: Exploring odds ratios, relative risks, and the logit transformation

## odds ratios

A function to compute the relative risk, given two conditional probabilities for the same event:

```{r, "TRY efficacy", purl=TRUE, results = "hide"}
relative_risk <- function(p0, p1) {
  return( p1 / p0 )
}
relative_risk(0.05, 0.15)
```

* Write a similar function to compute the odds ratio.
* Compare odds ratios to relative risks under a variety of scenarios

```{r}
odds <- function(p) { 
  return( p / (1-p))
  }
odds_ratio <- function(p0, p1) {
  return( odds(p1) / odds(p0) )
}
```


```{r}
relative_risk(0.05, 0.15)
odds_ratio(0.05, 0.15)
# the two summaries become more similar when the individual rates are near zero or one: 
relative_risk(0.03, 0.04)
odds_ratio(0.03, 0.04)
# and diverge when the rates are closer to 0.5 : 
relative_risk(0.3, 0.4)
odds_ratio(0.3, 0.4)
```


## logistic function

Suppose we have the following data

```{r, "datashell", purl=TRUE, results = "hide"}
nsubj <- 160
dat <- data.frame(
  BWT = rnorm(nsubj, 75, 5),
  CAVGSS = exp(rnorm(nsubj, log(1), 1)),
  AE01 = NA # AE indicator: 1 signifies AE ; 0 signifies no AE
)
head(dat)
```

Let's plot the probability of an AE for two linear predictors

 - logit(p) = 0 + CAVGSS
 - logit(p) = -2 + 3*CAVGSS

```{r, "TRY logistic", purl=TRUE, results = "hide"}
expit <- function(x) 1 / (1+exp(-x))
ggplot(dat) +
  geom_line(aes(x = CAVGSS, y = expit(CAVGSS))) +
  geom_line(aes(x = CAVGSS, y = expit(-2 + 3 * CAVGSS)), colour = "red", linetype = 2)
```

Exercise: Modify the linear predictor as a function of `CAVGSS` to:

* Make the intercept lower
* Make the E-R relationship shallower

```{r}
ggplot(dat) +
  geom_line(aes(x = CAVGSS, y = expit(CAVGSS))) +
  geom_line(aes(x = CAVGSS, y = expit(-2 + 3 * CAVGSS)), colour = "red", linetype = 2) + 
  geom_line(aes(x = CAVGSS, y = expit(-5 + 3 * CAVGSS)), colour = "red", linetype = 3) + 
  geom_line(aes(x = CAVGSS, y = expit(-2 + 1.5 * CAVGSS)), colour = "red", linetype = 4)
```


Now, suppose we've observed these data

```{r, "TRY link", purl=TRUE, results = "hide"}
dat$Pred <- expit(-2 + 0.75 * dat$CAVGSS)
dat$AE01 <- rbinom(nrow(dat), 1, dat$Pred)
ggplot(dat) +
  geom_line(aes(x = CAVGSS, y = Pred)) +
  geom_rug(data = filter(dat, AE01 == 0), aes(x = CAVGSS), sides = "b") +
  geom_rug(data = filter(dat, AE01 == 1), aes(x = CAVGSS), sides = "t")
```


Exercise: Generate values for `AE01` so that they conform to a probit-link, rather than a logit-link model.

Hints:

* The probit function $\Phi^{-1}$ is implemented in R as `qnorm()`.
* The inverse-probit function $\Phi$ is implemented in R as `pnorm()`.

```{r, "TRY link", purl=TRUE, results = "hide"}
dat$Pred <- pnorm(-2 + 0.75 * dat$CAVGSS)
dat$AE01 <- rbinom(nrow(dat), 1, dat$Pred)
ggplot(dat) +
  geom_line(aes(x = CAVGSS, y = Pred)) +
  geom_rug(data = filter(dat, AE01 == 0), aes(x = CAVGSS), sides = "b") +
  geom_rug(data = filter(dat, AE01 == 1), aes(x = CAVGSS), sides = "t")
```






