---
title: "Workbook Bayes01"
author: "Introduction to BRMS for exposure-response with binary endpoints"
date: "`r Sys.Date()`"
output:
  html_document:
#    css: docs/src/styles/styles.css
    number_sections: true
    theme: united
    toc: true
    toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(haven)
library(survival)
library(GGally)
library(binom)
library(texreg)
library(mgcv)
library(brms)
library(tidybayes)
library(modelr)

expit <- function(x) 1 / (1+exp(-x))
# note plogs = expit
```


```{r, "TRY toydata", purl=TRUE, results = "hide"}
load('../data/aedat.RDS')
source('preprocess_data.R')
```


# Fitting a model using brms

```{r, "TRY firststan", purl=TRUE, results = "hide"}
mod01_stan <- brm(formula = AE01 ~ CAVGSS + PTTYPE,  
                  data=aedat, 
                  family = bernoulli(link = "logit"),
                  warmup = 500, 
                  iter = 2000, 
                  chains = 4, 
                  inits = "random", 
                  cores = 4,
                  seed = 123)
```


***Exercise:***  

* Look at convergence diagnostics, do you think the Markov chains have converged?
    - Trace plots (`mcmc_plot(mod01_stan, type='trace')`)
    - Rhat (`mcmc_plot(mod01_stan, type='rhat')`)
    - density plots (`mcmc_plot(mod01_stan, type='dens_overlay')`)
    
* Look at the posterior summaries for parameters (`summary(mod01_stan)`).  Is there an effect of exposure?
    *  Think about how you might want to plot this effect.  We'll work on this later.

***Exercise:***

* Run a new model (mod02_stan) which adds `STUDYID`  to mod01_stan.  That is, the new model will use `formula = AE01 ~ CAVGSS + PTTYPE + STUDYID` 


```{r}
mod2_stan <- brm(formula = ,  
                  data=aedat, 
                  family = bernoulli(link = "logit"),
                  warmup = 500, 
                  iter = 2000, 
                  chains = 4, 
                  inits = "random", 
                  cores = 4,
                  seed = 123)
```

* Look at the convergence diagnostics.  How do these compare to mod01?
    * What do you think is causing this?
    * Try looking at a pairs plot of the posterior samples (`mcmc_plot(mod02_stan, type='pairs')`)


***Exercise:***

* Run a new model (mod03_stan) which modifies mod02_stan by adding normal prior distributions for the regression coefficients (i.e., class='b').  What might be reasonable prior standard deviations to make these only weakly informative?
    * Use the code below as a starting point
    * Look at the notes and/or the help file for set_prior ("?set_prior")

```{r}
mod03_stan <- brm(formula = ,  
                  data=aedat, 
                  family = bernoulli(link = "logit"),
                  warmup = 500, 
                  iter = 2000, 
                  chains = 4, 
                  inits = "random", 
                  cores = 4,
                  prior = set_prior(),
                  seed = 123)
```

* Look at the convergence diagnostics.  How do these compare to mod02?
    * What do you think is causing this?
    * Try looking at a pairs plot (`mcmc_plot(mod02_stan, type='pairs)`)



## reading R help

R help pages can be .... helpful :)

```{r, "TRY extractors", purl=TRUE, results = "hide"}
?brm # See especially the "See Also" section
?summary.brmsfit # See expecially the "Value" section
?fixef
?predict
?predict.brmsfit
```

Exercise:

* Print out a table of estimated coefficients and their standard errors.
* Print out 95% confidence intervals for the model parameters.
* Print out 90% confidence intervals for the model parameters.

```{r}
fixef(mod01_stan)
fixef(mod01_stan,probs = c(0.05, 0.95))
```

