---
title: "Workbook"
author: "Model Fitting"
date: "`r Sys.Date()`"
output:
  html_document:
    # css: docs/src/styles/styles.css
    number_sections: true
    theme: united
    toc: true
    toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(haven)
library(GGally)
library(binom)
library(texreg)
library(mgcv)
library(DHARMa)

expit <- function(x) 1 / (1+exp(-x))
```



# Workbook: Fitting a logistic regression model

```{r, purl=TRUE, results = "hide"}
# Preamble of data assembly done in previous workbooks

load('../data/aedat.RDS')
source('preprocess_data.R')
```

Let's start by fitting a model with effects of steady-state Cavg and patient type.  Note that patient type is a factor. and the reference level is "HV".

```{r}
levels(aedat$PTTYPE)
```


```{r, "TRY firstglm", purl=TRUE, results = "hide"}
mod01 <- glm(AE01 ~ CAVGSS + PTTYPE,
             family = binomial(link = "logit"),
             data = aedat,
             )
```

Exercise:

* Re-fit the model using a probit link (mod01_probit)

```{r, "TRY firstglm", purl=TRUE, results = "hide"}
mod01_probit <- glm(AE01 ~ CAVGSS + PTTYPE,
             family = binomial(link = "probit"),
             data = aedat,
             )
```

* Re-fit the model with STUDYID as an additional covariate (mod02)

```{r}
mod02 <- glm(AE01 ~ CAVGSS + PTTYPE + STUDYID, # having verified that STUDYID is a factor var!
             family = binomial(link = "logit"),
             data = aedat,
             )
```


## Reading R help

R help pages can be overwhelming.  Nonetheless, take a look at the following help files.

```{r, "TRY extractors", purl=TRUE, results = "hide"}
?glm # See especially the "See Also" section
?summary.glm # See expecially the "Value" section
?confint
?predict
?predict.glm
```

Exercise:

* Print out a table of estimated coefficients and their standard errors.
* Print out 95% confidence intervals for the model parameters.
* Print out 90% confidence intervals for the model parameters.

```{r}
summary(mod02) # or, to trim it down a bit :
summary(mod02)$coefficients 
```


```{r}
confint(mod02)
```


## Model intercept

```{r, "TRY intercept", purl=TRUE, results = "hide"}
partab01 <- cbind(coef(mod01), confint(mod01))
partab01
```

Exercise:

* Make the intercept more interpretable by applying the inverse of the link function.  What scale is this?
* What exposure and covariate settings does the intercept correspond to?

```{r}
# You can extract the intercept row like this:
partab01['(Intercept)',]
```


```{r}
expit(partab01['(Intercept)',]) # this puts it on the probability scale
```


```{r}
# or equivalently: 
plogis(partab01['(Intercept)',])
```


## Re-coding predictors

```{r, "TRY recoding", purl=TRUE, results = "hide"}
ref_bwt <- median(aedat$BWT) # will want reference value later, so save as variable!
dat_mod <- aedat %>% mutate(
  BWT_norm = BWT - ref_bwt,
  PTTYPE = factor(PTTYPE, levels = c("PT1", "PT2", "HV"))
  )
?relevel
```

Exercise: 

* Re-fit model `mod01` including CAVGSS, the re-coded PTTYPE and normalized body weight.  What exposure and covariate settings does the intercept corresponds to?

```{r}
mod03 <- glm(AE01 ~ CAVGSS + PTTYPE + BWT_norm,
             family = binomial(link = "logit"),
             data = dat_mod, 
             )
```

* How many fold higher are the odds of an AE for `PT1` versus `HV`?

```{r}
partab03 <- cbind(coef(mod01), confint(mod01))
exp(partab03['PTTYPEPT1', ])
```

Solution note: so, about 3.5-fold higher. 


