---
title: "Workbook"
author: "Visualizing binary data"
date: "`r Sys.Date()`"
output:
  html_document:
    # css: docs/src/styles/styles.css
    number_sections: true
    theme: united
    toc: true
    toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(haven)
library(GGally)
library(binom)
library(texreg)
library(mgcv)
library(DHARMa)

expit <- function(x) 1 / (1+exp(-x))
```



# Workbook: Visualizing binary data

```{r, "TRY toydata", purl=TRUE, results = "hide"}
load('../data/aedat.RDS')
str(aedat)
```

* This data set is comprised of a two week study (protocol A) and a six week study (protocol B)
* The data set includes patients of type `PT2`, a patient type that (let's say) has not been studied at higher dose levels.
* Suppose further that a dose level under consideration for a phase 2 study in `PT2` would have typical value exposures near 2.5 ug/mL.
* The objective of *this* analysis is to determine whether the AE rate will be "sufficiently low" in `PT2` at that exposure.


## EDA: Data key

```{r, "TRY datakey", purl=TRUE, results = "hide"}
aedat$AETOXGR <- factor(aedat$AETOXGR, 0:3, labels=c("None","Mild","Moderate","Severe"))

(aedat %>% group_by(USUBJID) %>% n_groups()) == nrow(aedat)

aedat %>% ungroup() %>% count(AETOXGR,AE01)

aedat %>% arrange(USUBJID, TTE) %>% group_by(USUBJID) %>% filter(n()>1)

# Create a variable for time-to-serious AE's, matching "AE01"
aedat <-
  aedat %>% group_by(USUBJID) %>%
  # End of study for patients without a severe event
  mutate(TTE_SEVERE = case_when(
    STUDYID=="PROTA" ~ 2,
    STUDYID=="PROTB" ~ 6
  ),
  # Time of severe event for those that had one
  TTE_SEVERE = ifelse(AETOXGR=="Severe", TTE, TTE_SEVERE)
  )
```


Exercise: Evaluate the preceding code and based on that answer the following:

* If we fit a model for `AE01` (severe events), is it reasonable to treat each row as independent? If not, what can be done to address this issue?
* If we fit a model for `AE01`, what are the implications of combining data from studies of two different durations?
* How can we assess whether the differing durations are likely to be problematic?



## Derived variables

Both for EDA and for model-checking, it's generally helpful to have quartiles of exposure:

```{r, "TRY varderive", purl=TRUE, results = "hide"}
dat_use <-
  aedat %>% arrange(USUBJID, TTE_SEVERE) %>% slice(1) %>%
  group_by(PBO) %>%
  mutate(Quartile = ifelse(PBO == "PBO", "PBO",
                           paste0("Q", ntile(CAVGSS, n = 4))))
```


## EDA: Pairs plots

```{r, "TRY covplot", purl=TRUE, results = "hide", width=10, height=10}
dat_eda <- dat_use %>% ungroup() %>%
  select(STUDYID, SEXTXT, PTTYPE, CAVGSS, BWT, Quartile) %>%
  mutate( # tell plot fn how to plot vars
   CAVGSS = as.numeric(CAVGSS),
   BWT = as.numeric(BWT)
  )

p <- ggpairs(dat_eda,
        axisLabels = "internal",
        columnLabels = c(rep("", 3), # 4 categorical variables
                         "Cavg", "Weight",  # continuous variables
                         "" # Quartile
                         )
        ) + theme_light()
p

ggsave("ggpairs.pdf", p, height=10, width=10)
```

Suppose we were planning to fit a model that included all of the following covariates:

* Sex
* Patient type
* Study
* Weight
* (as well as Cavg)

Should we expect this to be problematic?


## EDA: Categorical covariate plot

```{r}
aedat %>% 
  group_by(SEXTXT) %>% 
  summarise(x = sum(AE01), n=n()) %>% 
  mutate(phat = x / n,
         lcl = binom.confint(x,n,methods = "wilson")$lower,
         ucl = binom.confint(x,n,methods = "wilson")$upper) %>% 
  ggplot(aes(x=SEXTXT, y=phat)) + 
  geom_col() +
  geom_errorbar(aes(ymin=lcl, ymax=ucl), 
                width = 0.2) +
  labs(x='', y='Probability of a severe AE')
```


## EDA: E-R relationship

```{r, "TRY ereda", purl=TRUE, results = "hide"}
dat_eda <- dat_use %>% group_by(Quartile) %>% mutate(MedConc = median(CAVGSS))
p <- ggplot() +
  geom_rug(data = filter(dat_eda, AE01 == 0),
           aes(x = CAVGSS), sides = "b") +
  geom_rug(data = filter(dat_eda, AE01 == 1),
           aes(x = CAVGSS), sides = "t") +
  # NB: default smoother is problematic for 0-1
  # data. More on this later but live with it
  # for now.
  geom_smooth(data = dat_eda,
              aes(x = CAVGSS, y = AE01),
              method='gam', formula=y~s(x),
              method.args = list(family='binomial')) +
  stat_summary(
    data = dat_eda,
    aes(x = MedConc, y = AE01, group = MedConc),
    fun = function(y) sum(y) / length(y),
    geom = "point"
  ) +
  stat_summary(
    data = dat_eda,
    aes(x = MedConc, y = AE01),   
    # Wilson CI recommended by Agresti and Coull (2000) review paper
    fun.min = function(y) {
      binom.confint(sum(y),length(y),
                    methods = "wilson")$lower
      },
    fun.max = function(y) {
      binom.confint(sum(y), length(y),
                    methods = "wilson")$upper
      },
    geom = "errorbar"
  )
p
p + facet_wrap(~STUDYID) # Note exposure quartile not re-computed per-study. This is probably what you want.  
p + facet_wrap(~SEXTXT)
p + facet_wrap(~PTTYPE)
```

Exercise: List several key findings based on the EDA, and note any implications or expectations that relate to modeling these data.

```{r}

```

