---
title: "Simulate data"
output:
  html_document:
    df_print: paged
---

* Assume a hazard model of the form: $h(t) = \lambda_0 \cdot \left( 1 + \frac{\theta \cdot C(t)}{EC50 + C(t)} \right)$
* Assume 2\% event rate over 6 months on placebo and 30% event rate if $C(t) >> EC50$
* Implies $\lambda_0 = -\log(0.02)/180$ and $\theta = 1 - \log(0.30)/\log(0.02)$
* PK model: $V (L) = 3.5*(WT/70)*exp(\eta_V)$, $CL (L/day) = 0.5*(WT/70)^{0.75}*\exp(\eta_{CL})/24$

Study with doses of 0, 5, 10, 40 mg, IV bolus.  50 subjects per arm.  10% uniform drop-out; independent of AE hazard.



```{r}
library(tidyverse)
library(survival)
library(mrgsolve)
library(survminer)

theme_set(theme_bw())

set.seed(1234)

n_per_dose = 50
doses = c(0,5,10,20,40)
n_total = n_per_dose * length(doses)

## time units = days

dsim = tibble(ID = 1:n_total,
              dose = rep(doses, each=n_per_dose),
              BWT = rlnorm(n_total, meanlog = log(80), sdlog = 0.2),
              eta_V = rnorm(n_total,0,0.30),
              eta_CL = rnorm(n_total, 0, 0.30),
              eta_KA = rnorm(n_total, 0.30)) %>% 
  mutate(VC = 4.5 * (BWT/70) * exp(eta_V),
         CL = 0.5  * (BWT/70)^0.75 * exp(eta_CL) * 24,
         KA = (0.92*24) * exp(eta_KA),
         EC50 = 2, LAMBDA0 = -log(0.98)/180,
         EMAX = log(.50)/log(.98)-1,
         evid=1,
         amt=dose,
         time=0,
         ii=1,
         addl = 180,
         cmt=1,
         u_ae = runif(n_total))

```



```{r}
mod <- mread_cache(model = "Day1",project = 'model/mrgsolve/')
```
```{r}
dout = mod %>% 
  data_set(dsim) %>% 
  carry_out(dose,KA,CL,VC,LAMBDA0,u_ae) %>% 
  mrgsim_df(del=1/24) %>% 
  mutate(Surv = exp(-CHAZARD))
```

```{r}
dout %>% 
  filter(time <10) %>% 
  ggplot(aes(x=time, y=CENT, group=ID, col=as.factor(dose))) +
  geom_line() +
  facet_wrap(~dose)+
  theme_bw()
```


```{r}
dout %>% 
  filter(ID %in% seq(1,201,by=50)) %>% 
  ggplot(aes(x=time, y=Surv, group=ID, col=as.factor(dose))) +
  geom_line() +
  theme_bw()
```
```{r}
dae = dout %>% 
  filter(Surv > u_ae) %>% 
  group_by(ID) %>% 
  slice(n()) %>% 
  mutate(AE = as.numeric(time < 180))
```

```{r}
dae %>% group_by(dose) %>% summarise(p=mean(AE))
```


# Boxplots of CAVG_ss by dose group
CAVG_ss = dose/CL


# Plot AE prob vs CAVG_ss

# Plot time to AE by dose group

# Plot time to AE by CAVG_ss quartile





