---
title: "Simulate data"
output:
  html_document:
    df_print: paged
---

* Assume a hazard model of the form: $h(t) = \lambda_0 \cdot \left( 1 + \frac{\text{EMAX} \cdot C(t)}{EC50 + C(t)} \right)$
* Assume 2\% event rate over 6 months on placebo and 30% event rate if $C(t) >> EC50$
* Implies $\lambda_0 = -\log(0.02)/180$ and $\theta = 1 - \log(0.30)/\log(0.02)$
* PK model: 
  - $V (L) = 3.5*(WT/70)*exp(\eta_V)$
  - $CL (L/day) = 0.5*(WT/70)^{0.75}*\exp(\eta_{CL})/24$

Study with doses of 0, 5, 10, 20, 40 mg, IV bolus.  50 subjects per arm.  10% uniform drop-out; independent of AE hazard.

To-Do:
-------

* Check censoring - doesn't seem to be working based on TTE plots?
* Add shrinkage for 'observed' CL?
* Run exploratory logistic regression model
* "standard plot"
   * 
* VPC from fitted model
   * Simulate AE (0,1)
   * Smooth

```{r}
library(tidyverse)
library(survival)
library(mrgsolve)
library(survminer)
library(binom)
library(readr)

theme_set(theme_bw())

set.seed(1234)

n_per_dose = 50
doses = c(0,5,10,20,40)
n_total = n_per_dose * length(doses)


levels.dose = c("Placebo",paste0(doses[-1]," (mg)"))
levels.ecog = c("0","1",">1")
levels.smoke = c("Never Smoked","Current/Past Smoker")

## time units = days

dsim = tibble(ID = 1:n_total,
              dose = rep(doses, each=n_per_dose),
              dosef = factor(ifelse(dose>0,paste0(dose," (mg)"),"Placebo"),levels=levels.dose),
              BWT = rlnorm(n_total, meanlog = log(80), sdlog = 0.2),
              AGE = rnorm(n_total, 40, 5),
              cAGE = AGE-40,
              qAGE = cut(AGE,c(0,35,40,45,100),include.lowest = TRUE),
              eta_V = rnorm(n_total,0,0.30),
              eta_CL = rnorm(n_total, 0, 0.30),
              eta_KA = rnorm(n_total, 0.30),
              ECOGN = sample(c(0:2),size=n_total,replace=TRUE,prob=c(.50,.25,.25)),
              ECOGF = factor(case_when(ECOGN==0~levels.ecog[1],
                                       ECOGN==1~levels.ecog[2],
                                       ECOGN==2~levels.ecog[3]),levels=levels.ecog),
              ECOG1 = ifelse(ECOGN==1,1,0),
              ECOG2 = ifelse(ECOGN>1,1,0),
              TIMEDO = sample(c(0:1),size=n_total,replace=TRUE,prob=c(.90,.10)),
              SMOKEN = sample(c(0:1),size=250,replace=TRUE,prob=c(.60,.40)),
              SMOKEF = factor(ifelse(SMOKEN,levels.smoke[1],levels.smoke[2]),levels=levels.smoke)) %>% 
  # 10% of subjets drop out otherwise they have DO TIME of 180 days (~ administrative censoring)
  mutate(TIMEDO = ifelse(TIMEDO,runif(n_total,0,180),180)) 

dsim = dsim %>% 
  mutate(VC = 4.5 * (BWT/70) * exp(eta_V),
         CL = 0.5  * (BWT/70)^0.75 * exp(eta_CL) * 24,
         KA = (0.92*24) * exp(eta_KA),
         EC50 = 2, 
         LAMBDA0 = -log(0.98)/180,
         EMAX = log(.50)/log(.98)-1,
         beta1 = .22, # log(1.25) ~ 25% increase in hazard ECOG=1 vs ECOG=0,
         beta2 = .33, # log(1.4) ~ 40% increase in hazard ECOG>1 vs ECOG=0,
         beta3 = .095, # log(1.1) increase in hazard for age 
         evid=1,
         amt=dose,
         time=0,
         ii=1,
         addl = 180,
         cmt=1,
         u_ae = runif(n_total))


```


```{r}
mod <- mread_cache(model = "Day1",project = 'model/mrgsolve/')

dout = mod %>% 
  data_set(dsim) %>% 
  carry_out(dose,KA,CL,VC,LAMBDA0,u_ae,TIMEDO,ECOG1,ECOG2,cAGE) %>% 
  mrgsim_df(del=1/24) %>% 
  mutate(Surv = exp(-CHAZARD)) %>% 
  left_join(dsim %>% distinct(ID,ECOGF,SMOKEF,qAGE,AGE,dosef))

# lets simulate AE time
# subjects not observed to have an AE time are given an AE time of 180 days 
dout = dout %>% 
  group_split(ID) %>% 
  lapply(function(x) {
        x %>% mutate(TIMEAE=suppressWarnings(approx(x=1-x$Surv,y=x$time,xout=x$u_ae[1],yright=180)$y))
  }) %>% bind_rows() %>% 
  # taking into consideration censoring/drop out
  mutate(TIMEAE=ifelse(TIMEAE<TIMEDO,TIMEAE,TIMEDO),
         # indicator of whether subject had AE
         AE=ifelse(TIMEAE<TIMEDO,1,0))

dout = dout %>% 
    mutate(CAVGSS=dose/CL) 
dout = dout %>% 
  filter(CAVGSS>0) %>% 
  mutate(CAVGSSQ = paste0("Q", ntile(CAVGSS, n = 4))) %>%
  distinct(ID,CAVGSSQ) %>% 
  right_join(dout) %>% 
  mutate(CAVGSSQ=ifelse(CAVGSS==0,"Placebo",CAVGSSQ),
         CAVGSSQ=factor(CAVGSSQ,levels=c("Placebo",paste0("Q",1:4))))

# dataset with real values of concentration/tte
dout %>% write_csv("data/derived/datdout.csv")

dat.AE = dout %>% distinct(ID,dose,KA,CL,VC,ECOGF,AGE,qAGE,cAGE,SMOKEF,CAVGSS,CAVGSSQ,dosef,AE)
# adverse event dataset 
dat.AE %>% write_csv("data/derived/datAE.csv")

dat.TTE = dout %>% distinct(ID,dose,KA,CL,VC,ECOGF,AGE,qAGE,cAGE,SMOKEF,CAVGSS,CAVGSSQ,dosef,AE,TIMEAE,TIMEDO)
# TTE dataset 
dat.TTE %>% write_csv("data/derived/datTTE.csv")


```

# Plot of central compartment concetration by dose
```{r}
dout %>% 
  filter(time <10) %>% 
  ggplot(aes(x=time, y=CENT, group=ID, col=as.factor(dose))) +
  geom_line() +
  facet_wrap(~dose)+
  theme_bw()
```

# Plot of probabily of adverse event for random subject from each dose group
```{r}
dout %>% 
  filter(ID %in% seq(1,201,by=50)) %>% 
  ggplot(aes(x=time, y=Surv, group=ID, col=as.factor(dose))) +
  geom_line() +
  theme_bw()
```

# Boxplot of CAVGSS by dose group
```{r}
dat.AE %>% 
  ggplot() + 
  geom_boxplot(aes(y=CAVGSS,x=dosef)) + 
  labs(y=expression(C["ave,ss"]),
       x="Dose (mg)")
```

# Plot probability of AE (true and estimated) vs CAVGSS
```{r}
dout %>% 
  filter(time==180) %>% 
  ggplot() + 
  geom_smooth(aes(y=1-Surv,x=CAVGSS,color="True probability"),se=FALSE, 
              method='gam') +
     geom_rug(data = dat.AE %>% filter(AE == 0),
              aes(x = CAVGSS), sides = "b") +
     geom_rug(data = dat.AE %>% filter(AE == 1),
              aes(x = CAVGSS), sides = "t") +
     geom_smooth(data = dat.AE,
                 aes(x = CAVGSS, y = AE, color="Estimated probability from data"),
                 se=FALSE,
                 method = "gam",
                 method.args = list(family="binomial")) + 
  labs(x=expression(C["ave,ss"]),
       y="Probability of Adverse Event",
       color="Smoothed Probability") + 
  ylim(c(0,1))
```


# Plot time to AE by dose group

```{r}
sfit <- survfit(Surv(TIMEAE, AE) ~ dosef, data = dat.TTE)
p <- ggsurvplot(sfit,
                fun="event",
                data=dat.TTE,
                ggtheme=theme_bw(),
                conf.int = FALSE,
                legend.labs=levels.dose,
                legend.title="Doses",
                xlab = "Time (days)", 
                ylab = "Probability of Adverse Event",
                ylim=c(0,1))
p  
```

# Plot time to AE by CAVGSS quartile
```{r}
sfit <- survfit(Surv(TIMEAE, AE) ~ CAVGSSQ, data = dat.TTE)
p <- ggsurvplot(sfit,
                fun="event",
                data=dat.TTE,
                ggtheme=theme_bw(),
                conf.int = FALSE,
                legend.labs=levels(dat.TTE$CAVGSSQ),
                legend.title=expression("Quartiles of C"["ave,ss"]),
                xlab = "Time (days)", 
                ylab = "Probability of Adverse Event",
                ylim=c(0,1))
p  
```

# Plot time to AE by ECOG status

```{r}
sfit <- survfit(Surv(TIMEAE, AE) ~ ECOGF, data = dat.TTE)
p <- ggsurvplot(sfit,
                fun="event",
                data=dat.TTE,
                ggtheme=theme_bw(),
                conf.int = FALSE,
                legend.labs=levels.ecog,
                legend.title="ECOG Status",
                xlab = "Time (days)", 
                ylab = "Probability of Adverse Event",
                ylim=c(0,1))
p  
```

# Plot time to AE by AGE quartile
```{r}
sfit <- survfit(Surv(TIMEAE, AE) ~ qAGE, data = dat.TTE)
p <- ggsurvplot(sfit,
                fun="event",
                data=dat.TTE,
                ggtheme=theme_bw(),
                conf.int = FALSE,
                legend.labs=levels(dat.TTE$qAGE),
                legend.title=expression("Quartiles of AGE (years)"),
                xlab = "Time (days)", 
                ylab = "Probability of Adverse Event",
                ylim=c(0,1))
p  
```

# Plot time to AE by smoking status
```{r}

sfit <- survfit(Surv(TIMEAE, AE) ~ SMOKEF, data = dat.TTE)
p <- ggsurvplot(sfit,
                fun="event",
                data=dat.TTE,
                ggtheme=theme_bw(),
                conf.int = FALSE,
                legend.labs=levels.smoke,
                legend.title=expression("Smoking Status"),
                xlab = "Time (days)", 
                ylab = "Probability of Adverse Event",
                ylim=c(0,1))
p  
```

EDA plots of probability of AE by CAVGSS, age, ECOG status, and smoking status
```{r,EDA}
# EDA plots incidence of AE

ret = tibble(cov=c("AGE","CAVGSS"),
             lab=c("Age (years)","Concentration Average at Steady State")) 
ret = ret %>% mutate(plot=map2(cov,lab, function(.cov,.lab)  {
  ds.tmp = dat.AE %>% mutate(cov=get(.cov))
  ds.tmp1 = ds.tmp %>% 
    mutate(Quartile = paste0("Q", ntile(cov, n = 4))) %>%
    group_by(Quartile) %>% 
    summarize(MedConc = median(cov,na.rm=TRUE),
              mean=mean(AE,na.rm=TRUE),
              min=binom.confint(sum(AE,na.rm=TRUE),length(AE),methods = "wilson")$lower,
              max=binom.confint(sum(AE,na.rm=TRUE),length(AE),methods = "wilson")$upper
    ) 
  ggplot() +
    geom_rug(data = ds.tmp %>% filter(AE == 0),
             aes(x = cov), sides = "b") +
    geom_rug(data = ds.tmp %>% filter(AE == 1),
             aes(x = cov), sides = "t") +
    geom_smooth(data = ds.tmp,
                aes(x = cov, y = AE),
                method = "gam",
                method.args=list(family='binomial'),
                formula = y ~ s(x)) +
    geom_point(data=ds.tmp1,aes(MedConc,mean)) + 
    geom_errorbar(data=ds.tmp1,aes(x=MedConc,ymin=min,ymax=max)) + 
    ylim(0,1) +
    labs(y="Probability of Adverse Event",
         x=.lab) + 
    theme_bw()
}))
  
ret$plot

ret = tibble(cov=c("ECOGF","SMOKEF","qAGE"),
             lab=c("ECOG Status","Smoking Status","Age (years)")) 
ret = ret %>% mutate(plot=map2(cov,lab, function(.cov,.lab)  {

  ds.tmp = dat.AE %>% mutate(cov=get(.cov))
  ds.tmp = ds.tmp %>% 
    group_by(cov) %>%
    summarise(
      mean=mean(AE,na.rm=TRUE),
      min=binom.confint(sum(AE,na.rm=TRUE),length(AE),methods = "wilson")$lower,
      max=binom.confint(sum(AE,na.rm=TRUE),length(AE),methods = "wilson")$upper)
    
  ds.tmp %>% 
    ggplot() + 
    geom_point(aes(cov,mean)) + 
    geom_errorbar(aes(x=cov,ymin=min,ymax=max)) + 
    ylim(0,1) +
    labs(y="Probability of Adverse Event",
         x=.lab) + 
    theme_bw() 
}))

ret$plot


```


```{r}
fit = glm(AE ~ CAVGSS + ECOGF + cAGE + SMOKEF, data=dat.AE, family='binomial')
summary(fit )
```

```{r, VPCs}

# VPCs

ret = tibble(cov=c("ECOGF","SMOKEF","qAGE"),
             lab=c("ECOG Status","Smoking Status","Age (years)"))
ret = ret %>% mutate(fit = list(fit))

#VPCs
col.vpc = RColorBrewer::brewer.pal(7,"Set1")[1:2]
names(col.vpc) = c("observed","predicted")

set.seed(100)
nsim = 100
prop <- function(x) sum(x, na.rm = TRUE) / sum(!is.na(x))

stat_fn <- function(simy, ds_orig, cov) {
  dati <- ds_orig
  dati$AE <- simy
  dati = 
  suppressMessages(
    dati %>%
    group_by(cov, CAVGSSQ) %>%
    summarise(pAE = prop(AE))
  )
}

ret = ret %>% 
  mutate(plot=pmap(list(fit,cov,lab),function(.fit,.cov,.lab)  {


    ds.tmp = dat.AE %>% 
      mutate(cov = get(.cov),
             cov = factor(cov,labels=glue::glue("{.lab} : {levels(cov)}")))
    
    
    sim_dat <- simulate(.fit, nsim)
    lab.x = c("Quartiles of CAVGSS")
    
    obs_stat = 
      ds.tmp %>%
      group_by(cov, CAVGSSQ) %>%
      summarise(pAE = prop(AE)) 
    
    sim_stat <- apply(sim_dat, 2, stat_fn, ds_orig = ds.tmp, cov = .cov) %>% bind_rows()
    
    sim_stat_q <-
      suppressMessages(
        sim_stat %>%
          group_by(cov,CAVGSSQ) %>%
          summarize(LB=quantile(pAE, prob=.05),
                    MED=median(pAE),
                    UB=quantile(pAE, prob=.95))
      )
    comb_stat <- full_join(obs_stat, sim_stat_q)
    
    p = ggplot(data = comb_stat) +
      geom_pointrange(aes(x = CAVGSSQ, y = MED, ymin = LB, ymax = UB, colour = names(col.vpc)[2])) +
      geom_point(aes(x = CAVGSSQ, y = pAE, colour = names(col.vpc)[1]), size = 3, alpha = 0.5) +
      facet_wrap(~ cov) + 
      scale_color_manual("",values=col.vpc) +
      ylim(0,1) +
      labs(y="Probability of Adverse Event",
           x=lab.x) + 
      theme_bw()
    p
  }))


ret$plot



ret = tibble(cov=c("AGE","CAVGSS"),
             lab=c("Age","Concentration Average at Steady State")) 
ret = ret %>% mutate(fit = list(fit))

ret = ret %>% mutate(plot=pmap(list(fit,cov,lab), function(.fit,.cov,.lab)  {
  ds.tmp = dat.AE
  ds.tmp = ds.tmp %>% mutate(cov=get(.cov))

  obs_stat <-
    ds.tmp %>%
    ungroup %>%
    mutate(pAE=predict(mgcv::gam(AE ~ s(cov,k=4) , data=. , family=binomial()), type="response")) %>% 
    group_by(cov) %>%
    summarize(pAE=mean(pAE)) # Could use distinct, should be identical
  stat_fn <- function(simy, dat_orig ) {
    dati <- dat_orig
    dati$pAE <- simy
    dati$pAE <- predict(mgcv::gam(pAE ~ s(cov,k=4), data= dati, family=binomial()), type="response") 
    dati %>% group_by(cov) %>% summarize(pAE=mean(pAE))
  }
  sim_dat <- simulate(.fit, nsim)
  
  sim_stat <-
    data.frame(sim_dat) %>%
    select_all() %>%
    map(~stat_fn(.,ds.tmp)) %>%
    bind_rows
  sim_stat_q <-
    sim_stat %>%
    group_by(cov) %>%
    summarize(LB=quantile(pAE, prob=c(0.05)),
              MED=median(pAE),
              UB=quantile(pAE, prob=.95)
    )
  comb_stat <- full_join(obs_stat, sim_stat_q)
  ggplot(data = comb_stat) +
    geom_rug(data = ds.tmp %>% filter(AE == 0),
             aes(x = cov), sides = "b") +
    geom_rug(data = ds.tmp %>% filter(AE == 1),
             aes(x = cov), sides = "t") +
    geom_ribbon(aes(x = cov, ymin = LB, ymax = UB, fill = names(col.vpc)[2]), alpha=.25) +
    geom_line(aes(x = cov, y = MED, colour = names(col.vpc)[2]), size = 1, alpha = .50, linetype = "dashed") +
    geom_line(aes(x = cov, y = pAE, colour = names(col.vpc)[1]), size = 1) +
    scale_color_manual("",values=col.vpc) +
    scale_fill_manual("",values=col.vpc) +
    ylim(0,1) +
    xlim(range(obs_stat$cov)) +
      labs(y="Probability of Adverse Event",
         x=.lab) + 
    theme_bw() 
}))

ret$plot



ret = tibble(cov=c("AGE","CAVGSS"),
             lab=c("Age","Concentration Average at Steady State")) 
ret = ret %>% mutate(fit = list(fit))

ret = ret %>% mutate(plot=pmap(list(fit,cov,lab), function(.fit,.cov,.lab)  {
  
  ds.tmp = dat.AE
  ds.tmp = ds.tmp %>% mutate(cov=get(.cov))
  ds.tmp = ds.tmp %>% 
    mutate(Quartile = paste0("Q", ntile(cov, n = 5)))
  
  ds.tmp = ds.tmp %>% 
    group_by(Quartile) %>% 
    summarise(MedConc = median(cov,na.rm=TRUE)) %>% 
    distinct(Quartile,MedConc) %>% 
    right_join(ds.tmp)
  
  obs_stat <-
    ds.tmp %>%
    ungroup %>%
    mutate(pAE=predict(mgcv::gam(AE ~ s(cov,k=4) , data=. , family=binomial()), type="response")) %>% 
    group_by(cov) %>%
    summarize(pAE=mean(pAE)) # Could use distinct, should be identical
  stat_fn <- function(simy, dat_orig ) {
    dati <- dat_orig
    dati$pAE <- simy
    dati %>% group_by(MedConc) %>% summarize(pAE=mean(pAE))
  }
  sim_dat <- simulate(.fit, nsim)
  
  sim_stat <-
    data.frame(sim_dat) %>%
    select_all() %>%
    map(~stat_fn(.,ds.tmp)) %>%
    bind_rows
  sim_stat_q <-
    sim_stat %>%
    group_by(MedConc) %>%
    summarize(mean=mean(pAE,na.rm=TRUE),
              min=binom.confint(sum(pAE,na.rm=TRUE),length(pAE),methods = "wilson")$lower,
              max=binom.confint(sum(pAE,na.rm=TRUE),length(pAE),methods = "wilson")$upper)
      
  ggplot(data = obs_stat) +
    geom_rug(data = ds.tmp %>% filter(AE == 0),
             aes(x = cov), sides = "b") +
    geom_rug(data = ds.tmp %>% filter(AE == 1),
             aes(x = cov), sides = "t") +
    geom_errorbar(data = sim_stat_q,
                  aes(x = MedConc, min = min, max = max), alpha=.50) +
    geom_line(aes(x = cov, y = pAE, colour = names(col.vpc)[1]), size = 1) +
    scale_color_manual("",values=col.vpc) +
    ylim(0,1) +
    xlim(range(obs_stat$cov)) +
      labs(y="Probability of Adverse Event",
         x=.lab) + 
    theme_bw() 
}))

ret$plot

```




