---
title: "Repeated time-to-event and integrated hazard models"
output:
  slidy_presentation:
#    theme: cerulean
    fig_width: 9
    fig_height: 5
    font_adjustment: 3
    transition: none
    css: slidystyles.css
    footer: Metrum Research Group &copy 2021
    mathjax: local
    self_contained: false
    df_print: kable
    variant: markdown+fancy_lists
bibliography: Day4-references.bib
csl: statistical-science.csl
#nocite: | 
#    @2795, @2192, @3684
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE, echo=FALSE)

library(tidyverse)
library(stringr)
library(haven)
library(survival)
library(survminer)
library(survMisc)
library(texreg)
library(tidybayes)
library(bayesplot)
library(muhaz)
library(rstan)

set.seed(314159)
theme_set(theme_bw())
bayesplot::color_scheme_set("viridis")
``` 



# Key learning objectives for today

* Review questions from previous weeks

* Coding models which require integration
   - Hazards without closed-form integrals
   - Variables which vary continuously with time
  
* Repeated time-to-event models

# Why do we need to integrate when fitting TTE models?

* The likelihood in a TTE model depends on
  - Hazard at the time of an event: $h(t)$
  - Cumulative hazard up to the event/censoring time: $H(t) = \int_0^t h(s)~ds$
  
* Thus, we need to be able to integrate the hazard function

* For 'simple' TTE models, the hazard in integreble

* As we move away from simple TTE models, we may be working with hazard functions that don't have closed-form integrals

# Examples

* Continuously time-varying predictors
  - PK-driven TTE model
  - PD-driven TTE model

* Non-standard hazard models
  - The hazard function can take any form as long as $h(t) \geq 0$

* Cure-rate models
  - Will return to these later
  
# Specifying the models in Stan

* We will focus on one-dimensional integrals here
  - Integrate hazard only
  - Assuming the continuous-time covariate is available as a closed-form function of time

* More complex (joint) models are feasible in Stan using ODE solver
  - Could cover as a follow-up class
  
  
# One-dimensional integrals in Stan

* We will use the `integrate_1d` function in Stan

* Suppose we want to calculate $H(t) = \int_a^b h(y | \theta, x_r, x_i) ~ dy$

* Inputs to `integrate_1d`
  - function to integrate (must have a strict signature): $h(y~|~\theta, x_r, x_i)$
  - left limit: $a$
  - right limit: $b$
  - parameters: $\theta$
  - real-valued covariates: $x_r$
  - integer valued covariates: $x_i$

# Integrating a Weibull hazard

For example,

```
real cum_hazard[Nsubj];

for (i in 1:Nsubj) {
    cum_haz[i] = integrate_1d(weibull_hazard,   // function to integrate
                             0.0,               // integrate from 0
                             event_time[i],    // integrate to the event time
                             { alpha, gamma, beta1, beta2 }, // parameter array
                             {AGE[i]},  // array of real-valued data
                             {ECOG[i]}  // array of integer-valued data
                             );
}
```

# The `weibull_hazard` function

Must follow a strict function signature and be defined in the function block in a Stan program.

```
real weibull_hazard(real x,          // Function argument
                    real xc,         // Complement of function argument on the domain (not used in this class)
                    real[] theta,    // parameters
                    real[] x_r,      // data (real)
                    int[] x_i) {     // data (integer)

    real alpha = theta[1];
    real gamma = theta[2];
    real beta1 = theta[3];
    real beta2 = theta[4];
    
    real hazard = (alpha/gamma) * pow(x/gamma, alpha-1) * exp(beta1*x_r[1] + beta2*x_i[1]);

    return hazard;
}
```

# Example

* We'll work through an analysis using the simulated survival data from the last class. 

* **Models for change in tumor size, appearance of new lesions and survival probability in patients with advanced epithelial ovarian cancer** @Zecchin2016-rw
    - DDMORE repository submission IDs: DDMODEL00000217, DDMODEL00000218
    - Data simulated from these models

* Original study
    * Patients with platinum-sensitive recurrent ovarian cancer 
    * Randomly assigned to receive gemcitabine plus carboplatin (Cb+G) or carboplatin alone (Cb), every 21 days
    * Primary objective was to compare progression-free survival (PFS)

* We will analyze OS and the relationship between tumor changes and OS, using simulated data


```{r}
# OS data
d <- read_csv('../data/source/DDmodel0218_Simulated_OS.csv', na = c('.','-99'))

dos <- d %>% 
  filter(TIME>0) %>% 
    group_by(ID) %>% 
  mutate(meanGem = mean(AUC1),
         Group = if_else(meanGem > 0, "Cb+G", "Cb")) %>% 
  ungroup() %>% 
  filter(CMT==2, EVID==0)

dts <-  read_csv('../data/source/DDmodel0217_Simulated_SLD.csv', na = c('.','-99')) %>% 
  filter(!is.na(DV)) %>% 
  left_join(dos %>% select(ID,Group))
```

# OS by treatment group

```{r, out.width='90%'}
ggsurvplot(survfit(Surv(TIME,DV)~Group, data=dos), palette = 'nejm', xlab='Time (days)', risk.table = TRUE)
```

# Tumor growth by treatment group

```{r, out.width='90%'}
dts %>% 
 ggplot(aes(x=TIME, y=DV, group=ID, col=Group)) +
  geom_line(alpha=0.3) +
  facet_wrap(~Group) +
  labs(x='Time (days)', y='Tumor size (mm)')
```

# Modeling approach 

* To keep the example manageable, we assume that the following tumor dynamic model has been fitted

\[
\frac{d}{dt} TS(t) = \left(KG_i - (KD0_{i} \cdot \overline{AUC0}_i + KD1_{i} \cdot \overline{AUC1}_i \right) TS(t)
\]

where $AUC0_i$ and $AUC1_i$ are model-predicted time-averaged AUC values for carboplatin and gemcitabine, respectively

* We use the post-hoc estimated parameters for $KG_i$, $KD0_i$, $KD1_i$ and $IBASE_i$.

# Modeling approach 
We'll include effects for the model-predicted baseline tumor size and the tumor size relative to baseline (RTS).

$$
h_i(t) = \alpha / \lambda_0  \times (t / \lambda_0)^{\alpha-1} \exp(\beta_1 \times RTS(t) + \beta_2 BTS_i)
$$

where $TS_i(t)$ comes from the tumor model predictions and

$$
RTS_i(t) = \frac{TS_i(t)}{BTS_i} 
$$

# Workbook

We'll try a few models in Workbook_time_varying.Rmd



# Models for repeated times to event (RTTE)

* Also known as recurrent event data

* Examples
  - Time to sleep onset in a sleep study
  - Time to falls in elderly subjects
  - Time to ear infection in children
  - Time to asthma exacerbation
  - Time to hospital admission in schizophrenia patients

# Time to bleeding event in patients with hemophilia A

Data were simulated from a PK-RTTE model by Abrantes et. al [@Abrantes2020-yd]

* PK model for a full-length recombinant human FVIII product
  * Dosed at levels of 20 - 50 IU/kg
  
* Repeated time to bleeding event model
  * Gompertz TTE model with an inhibitory Emax effect of FVIII(t)
  * Follow-up for one year
  
Simulated data:

* Hypothetical trial in 100 patients
* Randomly allocated 1:1 to doses of 20 and 50 IU/kg
* Follow-up for 90 days


```{r}
# Time to bleeding data
d <- read_csv('../data/derived/rtte_example.csv', na = c('.'))

d <- d %>% 
  group_by(ID) %>% 
  mutate(TIME = TIME/24,  # Convert from hours to days
         PREVTIME = lag(TIME, default = 0))

# Extract one record per subject.  The last record will indicate the number of bleeding events.
d_ind = d %>% group_by(ID) %>% slice_tail() %>% ungroup() %>% arrange(ID)

```


# Number of bleeding events

Number of events stratified by dose.   

```{r, out.width='80%'}
d_ind %>% 
  ggplot(aes(x=cumulative_events)) + 
  geom_bar() + 
  facet_grid(~DOSE)
```

# Time to bleeding 

Time to first, second, and third bleeding events.

```{r, out.width='80%'}
d_first = d %>% group_by(ID) %>% filter(TIME == min(TIME)) %>% ungroup()

d_second = d %>% 
  group_by(ID) %>% 
  filter((cumulative_events==2 & event==1) | (cumulative_events < 2 & TIME == max(TIME) & event==0)) %>% 
  ungroup()

d_third = d %>% 
  group_by(ID) %>% 
  filter((cumulative_events==3 & event==1) | (cumulative_events < 3 & TIME == max(TIME) & event==0)) %>% 
  ungroup()

fit1 <- survfit(Surv(TIME,event)~1, data = d_first)
fit2 <- survfit(Surv(TIME,event)~1, data = d_second)
fit3 <- survfit(Surv(TIME,event)~1, data = d_third)

ggsurvplot_combine(list(First=fit1, Second=fit2, Third=fit3))
```

# Key feature of RTTE data: correlation within subjects

* To model this correlation, we typically include a random subject effect.  For example,

$$
h_i(t) = h_0(t) \exp(x_i \beta) \times Z_i
$$
where
$$
Z_i \in (0,+\infty) \sim G(z ~ | ~ \omega) 
$$

* Common distributions
  * $Z_i \sim \text{log-normal}(0,\omega)$ (common in pharmacometrics)
     * Median = 1, mean = $\exp(\omega^2/2)$
  * $Z_i \sim \text{Gamma}(\omega,\omega)$ (common in statistics)
     * Mean = 1
  * In practice, difficult to differentiate frailty distributions [@Balan2020-wv] 
  
# Other uses for frailties

Frailties are also used for modeling other correlations/clustering

* Within centers in a multi-center trial
* Shared random effects models for joint longitudinal-TTE models

# Model fitting

* Need to carefully consider if and how the hazard changes fare an event
  * Does the hazard completely reset after an event?
  * Does the hazard depend on the number of events preceding $t$?
  
* Is the subject at-risk for another event immediately after experiencing an event?
    - Subjects may be  out of risk set if 'event' has a period of time 
      - time to sleep onset (not at 'risk' again until awake)
      - time to hospitalization (not at risk while in the hospital)
  - If event is instantaneous, then not out of risk set
  - Use multi-state models, if interest is in time in each state


# Likelihood function

* The likelihood depends on
  - Hazard at the time of an event
  - Accumulated hazard between events
  
* Conditional on the random effect, events are assumed to be independent

# Likelihood function example

* Suppose a subject experienced events at times $t_1$ and $t_2$ and is event-free until $t_3$ when she is out of the risk set

* The likelihood contribution is

$$
\begin{align*}
L_i(\theta) &= \int_0^\infty \left( h_i(t_1~|~Z) ~S_i(t_1~|~Z) \right) \times \left( h_i(t_2~|~Z) \frac{S_i(t_2~|~Z)}{S_i(t_1~|~Z)} \right) \times \frac{S_i(t_3~|~Z)}{S_i(t_2~|~Z)} f(Z) dZ \\
  &= \int_0^\infty h_i(t_1~|~Z) \times h_i(t_2~|~Z) \times S_i(t_3~|~Z) ~ f(Z) ~ dZ
\end{align*}
$$
Conditional on $Z$, the log-likelihood contribution would be

$$
\begin{align*}
\ell_i(\theta) &= \log h_i(t_1~|~Z) - H_i(t_1~|~Z) + \log h_i(t_2~|~Z) - \left(H_i(t_2~|~Z) - H_i(t_1~|~Z) \right) - \left( H_i(t_3~|~Z) - H_i(t_2~|~Z) \right) \\
  &= \log h_i(t_1~|~Z) + \log h_i(t_2~|~Z) - H_i(t_3~|~Z)
\end{align*}
$$

# Bayesian analysis of RTTE models

* Similar to previous modeling, but with the added random effect distribution

* Prior distribution on $\omega$

* We'll model using log-normal distributions for $Z$, but try using a Gamma model for homework

# Simulating from an RTTE model

* Much more complicated than single TTE simulation
  * Simulate $T1$ - time to first event
  * Simulate $T2$ - time to second event, given first event
  * Etc.
  
* Inter-event intervals depend on accumulated hazard following previous event
  * Easiest with closed-form cumulative hazard functions
  
* Use inverse CDF method for simulating TTE
  - $T_1 = S^{-1}(u_1)$ where $u_1 \sim U(0,1)$
  - $T_2 = T_1 + \bar{S}^{-1}(u_2)$ where $\bar{S}(u) = \frac{S(u)}{S(T1)}$ and $u_2 \sim U(0,1)$
  - Etc.
  - Simulate until $T_k > \text{follow-up time}$

* For more details see [@Jahn-Eimermacher2015-rz]

# Workbook_rtte

We'll work through these using a workbook.


# References