---
title: "RTTE Notes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

RTTE TOPICS

* Inclusion of random effects / frailties

* log-likelihood depends on cumulative hazard between events
  - first event = $\delta_i \log h(t1_i) -H(t1)$ = $\delta_i \log h(t1_i) - \int_0^{t1_i} h(s) ds$
  - second event = $\delta_i \log h(t2_i) -(H(t2_i) - H(t1))$ = $\delta_i \log h(t2) - \int_{t1}^{t2} h(s) ds$
* 
* Does the hazard re-set after the event?
  - Insert picture of hazard not/re-setting
  - Examples Time to sleep

* Risk set
  - Subjects may be  out of risk set if 'event' has a period of time 
      - time to sleep onset (not at 'risk' again until awake)
      - time relapse in MS
  - If event is instantaneous, then not out of risk set

* Multi-state models if interest falls on both 'events'/states


Simulating from time to bleeding model in
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7193498/


```{r}
library(mrgsolve)
library(tidyverse)
library(survival)
library(survminer)
```

```{r}
# mod <- modlib(model='pk2cmt')
mod <- mread('../model/mrgsolve/rtte_sim.cpp')
# dose = 20 IU/kg Q3D (50 IU/kg Q3D)
set.seed(9642)
idata <- data.frame(ID=1:100) %>% 
  mutate(BWT = rnorm(n(), 76, 13),
         TIME=0,
         AMT=rep(c(20,50),each=50)*BWT,  # IU
         CMT='CENT',
         EVID=1,
         ADDL = 30,  # 90 days of dosing
         II = 3 * 24,  # 3 days
         CL = 1.93 * exp(rnorm(n(),0,.307)),  # Table 
         VC = 30.3 * exp(rnorm(n(),0,.151)),
         Q = 1.69,
         VP = 6.29,
         lambda = exp(rnorm(n(), 2.96, 1.36))
         )

```

```{r}
sims = mod %>% data_set(idata) %>% carry_out(CL) %>% mrgsim(tgrid = 1:(24*90))
```

```{r}
sims %>% as.data.frame() %>% 
  filter(TIME<24*7) %>% 
  ggplot(aes(x=TIME, y=exp(-hazard), group=ID)) +
  geom_line()
```

```{r}
sims %>% as.data.frame() %>% 
#  filter(TIME<24*7) %>% 
  group_by(TIME) %>% summarise(p = mean(exp(-hazard))) %>% 
  ggplot(aes(x=TIME, y=p)) +
  geom_line()
```

```{r}
sims <- sims %>% 
  as.data.frame() %>% 
  left_join(idata %>% 
              select(ID,AMT,BWT) %>% 
              # Induce 30% censoring
              mutate(censoring_time = if_else(runif(n())<0.30, 90*24*runif(n()), 90*24))
            ) %>% 
  mutate(u = runif(n())) %>% 
  filter(TIME <= censoring_time) %>% 
  group_by(ID) %>% 
  mutate(prev_hazard = lag(hazard, default = 0),
         event = as.numeric(u < (hazard - prev_hazard)),
         cumulative_events = cumsum(event)) %>% 
  # Truncate data at 25th event
  filter(cumulative_events < 25 | (cumulative_events==25 & event==TRUE)) %>% 
  # Filter to event and censoring record
  filter(event==TRUE | event==F & TIME==max(TIME)) %>% 
  ungroup() %>% 
  # Keep only relevant variables for modeling
  select(ID, BWT, AMT, CL, TIME, event, cumulative_events) %>% 
  mutate(CAVGSS = AMT/CL/(24*3),
         DOSE = AMT/BWT) %>% 
  select(-AMT)
```
Save simulated data

```{r}
sims %>% 
  mutate(CL = signif(CL,3),
         BWT = signif(BWT,3),
         CAVGSS = signif(CAVGSS, 3),
         DOSE = round(DOSE)) %>% 
  write_csv(file='../data/derived/rtte_example.csv', na='.')
```

```{r}
sims %>% group_by(ID) %>% summarise(n=sum(event)) %>% count(n)
```

Time to first bleeding event

```{r}
d1 = sims %>% 
  group_by(ID,DOSE) %>% 
  slice(1)

fit1 <- survfit(Surv(TIME,event)~1, data=d1)

```

Time to second bleeding event

```{r}
d2 = sims %>% 
  filter(cumulative_events <= 2) %>% 
  group_by(ID,DOSE) %>%
  summarise(TIME=max(TIME),
            event2 = max(cumulative_events)==2)

fit2 <- survfit(Surv(TIME,event2)~1, data=d2)

```
Time to third bleeding event

```{r}
d3 = sims %>% 
  filter(cumulative_events <= 3) %>% 
  group_by(ID,DOSE) %>%
  summarise(TIME=max(TIME),
            event3 = max(cumulative_events)==3)

fit3 <- survfit(Surv(TIME,event3)~1, data=d3)

```

Plot time to first, second, and third bleeding events

```{r}
ggsurvplot_combine(list(First=fit1, Second=fit2, Third=fit3))
```

