---
title: "Workbook for Introduction to TTE modeling"
author: "Integrating the hazard function in Stan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(survival)
library(survminer)
library(texreg)
library(mgcv)
library(muhaz)
library(rstan)
library(bayesplot)
library(tidybayes)

theme_set(theme_bw())
bayesplot::color_scheme_set("viridis")

```


```{r}
# OS data
d <- read_csv('../data/source/DDmodel0218_Simulated_OS.csv', na = c('.','-99'))  %>% 
  group_by(ID) %>% 
  mutate(pTIME = lag(TIME,default=0)) 

# Pull out the on-treatment records
d_on_trt = d %>% 
  filter(EVID==2 & (AUC0>0 | AUC1 > 0)) 

# Pull out the off-treatment records
d_off_trt = d %>% 
  filter(TIME>0 & CMT==2)

# Calculate time-weighted AYC values for driving the model
auc_means = d_on_trt %>% 
  group_by(ID) %>% 
  summarise(auc0bar = sum(AUC0*(TIME-pTIME))/sum(TIME-pTIME), 
            auc1bar = sum(AUC1*(TIME-pTIME))/sum(TIME-pTIME))

# Combine data
model_data = d_on_trt %>% group_by(ID) %>% slice_tail() %>%
  bind_rows(d_off_trt) %>% 
  left_join(auc_means) %>% 
  arrange(ID,TIME) %>% 
  group_by(ID) %>% 
  mutate(pTIME = lag(TIME,default=0.04),
         mauc0 = if_else(!is.na(CMT),AUC0,auc0bar),
         mauc1 = if_else(!is.na(CMT),AUC1,auc1bar),
         TSBASE = IBASE*1000) %>% 
  select(ID,TIME,pTIME,mauc0,mauc1,KG,KD0,KD1,TSBASE)

new_lesion_data <- d %>% 
  filter(NWLSCOV==1) %>% 
  group_by(ID) %>%
  summarise(NLIND = 1,
            NLTIME = min(TIME))

model_data <- model_data %>% 
  left_join(new_lesion_data) %>% 
  ungroup() %>% 
  mutate(NLIND = if_else(is.na(NLIND), 0, NLIND),
         NLTIME = if_else(is.na(NLTIME), 10000, NLTIME),
         row = 1:n())

model_data_subject <- d %>% filter(CMT==2, TIME>0) %>% 
  select(ID, event=DV, event_time=TIME) %>% 
  # Add first and last row values 
  left_join(model_data %>% group_by(ID) %>% 
              summarise(row_first = min(row),
                        row_last = max(row))) %>% 
  # Add last observation time
  left_join(d_on_trt %>% 
              arrange(ID,TIME) %>% 
              group_by(ID) %>% 
              slice_tail() %>% 
              select(ID,last_obs_time=TIME))
```

For faster run times, lets cut down to 100 subjects

```{r}
model_data <- model_data %>% filter(ID <= 100)
model_data_subject <- model_data_subject %>% filter(ID <= 100)
```

```{r}
stan_data = list(
  N = nrow(model_data),
  Nsubj = nrow(model_data_subject),
  
  ID = model_data$ID,
  TIME = model_data$TIME,
  PREVTIME = model_data$pTIME,
  AUC0 = model_data$mauc0,
  AUC1 = model_data$mauc1,
  KG = model_data$KG,
  KD0 = model_data$KD0,
  KD1 = model_data$KD1,
  TSBASE = model_data$TSBASE,
  NLTIME = model_data$NLTIME,
  NLIND = model_data$NLIND,
  
  IDsubj = model_data_subject$ID,
  event = model_data_subject$event,
  event_time = model_data_subject$event_time,
  row_first = model_data_subject$row_first,
  row_last = model_data_subject$row_last,
  last_obs_time = model_data_subject$last_obs_time
)
```


# Working with traditional statistics parameterization of Weibull model


\begin{align*}
h_0(t) &= \alpha / \lambda_0  \times (t / \lambda0)^{\alpha-1}\\
\\
H_i(t) &= ( t / \lambda0 )^\alpha \\
S_i(t) &= \exp(-(t/\lambda0)^\alpha)
\end{align*}


Under this parameterization, 

* median OS = $\left( \frac{\log 2}{\lambda_0} \right)^{1/\alpha}$
* mean OS = $ \lambda_0^{-1/\alpha} ~~\Gamma(1 + 1/\alpha)$



```{r, results='hide'}
fit_weibull_base <- stan(file = '../model/stan/Day4/timevarying_weibull_base.stan',
                        data = stan_data, 
                        chains = 4,
                        iter=800,
                        cores = 4,
                        seed = 76341
                        )
```


```{r}
print(fit_weibull_base, pars=c('log_lambda0','log_shape'))
```


## Adding effect of tumor size to the model

For the sake of simplicity, let's assume that we're fitting this model in a two-stage approach, and that we already have estimates from the tumor modeling.

The tumor model is of the form:

\[
\frac{d}{dt} TS(t) = \left(KG_i - (KD0_{i} \cdot \overline{AUC0}_i + KD1_{i} \cdot \overline{AUC1}_i \right) TS(t)
\]
 where $AUC0_i$ and $AUC1_i$ are model-predicted time-averaged AUC values for carboplatin and gemcitabine, respectively.


We'll include effects for the model-predicted baseline tumor size and the tumor size relative to baseline (RTS).



$$
h_i(t) = \alpha / \lambda_0  \times (t / \lambda_0)^{\alpha-1} \exp\left(\beta_1 \times \log RTS(t) + \beta_2 \times \log (BTS_i - \overline{BTS}) \right)
$$

where $TS_i(t)$ comes from the tumor model predictions and

$$
RTS_i(t) = \frac{TS_i(t)}{BTS_i} 
$$


```{r, results='hide'}
fit_weibull_ts <- stan(file = '../model/stan/Day4/timevarying_weibull_nl.stan',
                        data = stan_data, 
                        chains = 4,
                        iter=800,
                        cores = 4,
                        seed = 76341
                        )
```

```{r}
print(fit_weibull_ts, pars=c('log_lambda0','log_shape','beta1','beta2'))
```

```{r}
samples_int <- spread_draws(fit_weibull_ts, beta1,beta2, log_lambda0,log_shape) %>% 
  rename(Chain=.chain)
```

```{r}
mcmc_trace(samples_int, pars=c('beta1','beta2','log_lambda0','log_shape'))
```

```{r}
mcmc_pairs(samples_int, pars=c('beta1','beta2','log_lambda0','log_shape'))
```

Let's look at and compare the model fits

```{r}
loo_base = loo(fit_weibull_base,pars='log_like', save_psis=TRUE)
loo_base
```

```{r}
loo_ts = loo(fit_weibull_ts,pars='log_like', save_psis=TRUE)
loo_ts
```

```{r}
loo::loo_compare(loo_base,loo_ts)
```

The `loo` output for the covariate model indicates that the model fits some subjects poorly.  Let's investigate:

```{r}
ids_to_check = which(loo_ts$diagnostics$pareto_k > 0.5)
loo_ts$diagnostics$pareto_k[ids_to_check]
```

```{r}
model_data %>% filter(ID %in% ids_to_check)
```


