---
title: "Workbook for Introduction to TTE modeling"
author: "Integrating the hazard funtion in Stan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
params:
  include: TRUE
editor_options: 
  chunk_output_type: console
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(survival)
library(survminer)
library(texreg)
library(mgcv)
library(muhaz)
library(rstan)
library(bayesplot)
library(tidybayes)

theme_set(theme_bw())
bayesplot::color_scheme_set("viridis")

```


```{r}
# OS data
d <- read_csv('../data/source/DDmodel0218_Simulated_OS.csv', na = c('.','-99'))

# Add week 12 (Day 84) predicted tumor size
d84 <- d %>% 
  filter(TIME <= 84) %>% 
  group_by(ID) %>% 
  mutate(rate = KG/1000 - KD0/1000*AUC0 - KD1/100*AUC1,
         prevTIME = lag(TIME,default = 0),
         change = exp(rate * (TIME-prevTIME)),
         ipred = IBASE * 1000 * cumprod(change)
         ) %>% 
  arrange(ID,TIME) %>% 
  slice(n()) %>% 
  mutate(ipred84 = ipred * exp(rate * (84-TIME)),
         rts84 = ipred84 / ( IBASE * 1000 ) )

dos <- d %>% 
  filter(TIME>0) %>% 
  group_by(ID) %>% 
  mutate(meanGem = mean(AUC1),
         Group = if_else(meanGem > 0, "Cb+G", "Cb")) %>% 
  ungroup() %>% 
  filter(CMT==2, EVID==0) %>% 
  left_join(d84 %>% select(ID, ipred84, rts84)) %>% 
  mutate(rts84_f = paste0("Q", ntile(rts84, n = 4)))

dos84 <- dos %>% 
  filter(TIME>84) %>% 
  mutate(TIME = TIME-84)


```


# Working with traditional pharmacometric parameterization of Weibull model


\begin{align*}
h_i(t) &= h_0(t) \times \exp(\beta \times \text{RTS}_i)  \\
h_0(t) &= \lambda_0 \times \alpha \ \times t^{\alpha-1}\\
\\
H_i(t) &= \lambda_i t^\alpha \\
S_i(t) &= \exp(-\lambda_i t^\alpha)
\end{align*}


Under this parameterization, 

* median OS = $\left( \frac{\log 2}{\lambda_0 \exp(\beta \times RTS)} \right)^{1/\alpha}$
* mean OS = $\left\{ \lambda_0 \exp(\beta \times RTS) \right\}^{-1/\alpha} ~~\Gamma(1 + 1/\alpha)$


Set-up the data

```{r}
tte_emax_data <- list(
  N = nrow(dos84),
  Y = dos84$TIME/365,  # Use years instead of days
  cens = 1-dos84$DV,
  K = 1,
  X = model.matrix(~-1 + ECOG, data=dos84),
  RTS = dos84$rts84,
  prior_only = 0
)
```



```{r, results='hide'}
fit_weibull_int <- stan(file = '../model/stan/weibull_integrated_hazard.stan',
                        data = tte_emax_data, 
                        chains = 4,
                        iter=1000,
                        cores = 4,
                        seed = 76341
                        )
```

```{r}
print(fit_weibull_int, pars=c('slope','lambda0','b','shape'))
```

```{r}
samples_int <- spread_draws(fit_weibull_int, slope,b[i],lambda0,shape) %>% 
  rename(Chain=.chain)
```

```{r}
mcmc_trace(samples_int, pars=c('slope','b','lambda0','shape'))
```

```{r}
mcmc_pairs(samples_int, pars=c('slope','b','lambda0','shape'))
```

** Exercise: **

How does this compare to the closed-form model?  Re-fit the closed-form model and compare parameter estimates.

Try coding a model with a 'hockey stick' hazard:

$$
h(t) = \begin{cases} \alpha_1 + \beta_1 t & 0 \leq t < \tau_1 \\
                     \alpha_1 + \beta_1 \tau_1 + \beta_2 (t-\tau_1) & t \geq \tau_1
\end{cases}
$$
where you can either pick a fixed value of $\tau_1$ or estimate it.

How does this compare to the Weibull model fit?


