---
title: "Workbook for Introduction to TTE modeling"
author: "Integrating the hazard funtion in Stan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(survival)
library(survminer)
library(texreg)
library(mgcv)
library(muhaz)
library(rstan)
library(bayesplot)
library(tidybayes)

theme_set(theme_bw())
bayesplot::color_scheme_set("viridis")

```


```{r}
# OS data
d <- read_csv('../data/source/DDmodel0218_Simulated_OS.csv', na = c('.','-99'))

# Add week 12 (Day 84) predicted tumor size
d84 <- d %>% 
  filter(TIME <= 84) %>% 
  group_by(ID) %>% 
  mutate(rate = KG/1000 - KD0/1000*AUC0 - KD1/100*AUC1,
         prevTIME = lag(TIME,default = 0),
         change = exp(rate * (TIME-prevTIME)),
         ipred = IBASE * 1000 * cumprod(change)
         ) %>% 
  arrange(ID,TIME) %>% 
  slice(n()) %>% 
  mutate(ipred84 = ipred * exp(rate * (84-TIME)),
         rts84 = ipred84 / ( IBASE * 1000 ) )

dos <- d %>% 
  filter(TIME>0) %>% 
  group_by(ID) %>% 
  mutate(meanGem = mean(AUC1),
         Group = if_else(meanGem > 0, "Cb+G", "Cb")) %>% 
  ungroup() %>% 
  filter(CMT==2, EVID==0) %>% 
  left_join(d84 %>% select(ID, ipred84, rts84)) %>% 
  mutate(rts84_f = paste0("Q", ntile(rts84, n = 4)))

dos84 <- dos %>% 
  filter(TIME>84) %>% 
  mutate(TIME = TIME-84)


```


# Working with traditional pharmacometric parameterization of Weibull model


$$
\begin{align*}
h_i(t) &= h_0(t) \times \exp(\beta \times \text{RTS}_i)  \\
h_0(t) &= \lambda_0 \times \alpha \ \times t^{\alpha-1}\\
\\
H_i(t) &= \lambda_i t^\alpha \\
S_i(t) &= \exp(-\lambda_i t^\alpha)
\end{align*}
$$
Under this parameterization, 

* median OS = $\left( \frac{\log 2}{\lambda_0 \exp(\beta \times RTS)} \right)^{1/\alpha}$
* mean OS = $\left\{ \lambda_0 \exp(\beta \times RTS) \right\}^{-1/\alpha} ~~\Gamma(1 + 1/\alpha)$


Set-up the data

```{r}
tte_emax_data <- list(
  N = nrow(dos84),
  Y = dos84$TIME/365,  # Use years instead of days
  cens = 1-dos84$DV,
  K = 1,
  X = model.matrix(~-1 + ECOG, data=dos84),
  RTS = dos84$rts84,
  prior_only = 0
)
```



```{r, results='hide'}
fit_weibull_int <- stan(file = '../model/stan/weibull_integrated_hazard.stan',
                        data = tte_emax_data, 
                        chains = 4,
                        cores = 2,
                        seed = 76341
                        )
```

```{r}
print(fit_weibull_int, pars=c('slope','lambda0','b','shape'))
```

```{r}
samples_int <- spread_draws(fit_weibull_int, slope,b[i],lambda0,shape) %>% 
  rename(Chain=.chain)
```

```{r}
mcmc_trace(samples_emax, pars=c('slope','b','lambda0','shape'))
```

```{r}
mcmc_pairs(samples_emax, pars=c('slope','b','lambda0','shape'))
```

Compare models using loo


```{r}
loo_pmx <- loo::loo(extract(fit_weibull_pmx, pars='log_lik')$log_lik)
loo_emax <- loo::loo(extract(fit_weibull_emax, pars='log_lik')$log_lik)

loo_compare(loo_pmx, loo_emax)

```
Plot estimated relationship with RTS

```{r}
RTS_grid <- seq(min(dos84$rts84), max(dos84$rts84), length=100)

samples_emax %>% crossing(rts = RTS_grid) %>% 
  mutate(log_hr = emax*rts / (e50+rts)) %>% 
  group_by(rts) %>% 
  summarise(est = median(log_hr), 
            lcl=quantile(log_hr,probs=0.05),
            ucl = quantile(log_hr, probs = 0.95)) %>% 
  ggplot(aes(x=rts, y=est)) +
  geom_line() +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.2) +
  labs(x='Relative change in tumor size to week 12', y='log hazard ratio') 
```

Simple VPCs without censoring

```{r}
os_sims_emax <- gather_draws(fit_weibull_emax, Ysim[ID]) %>% 
  left_join(dos84)
head(os_sims_emax)
```

Survival VPC by ECOG status 

```{r, echo=TRUE, out.height='50%'}
# Apply to observed data
obs_surv = vpc_stat_km(dos84 %>% mutate(time=TIME/365, event=DV), group='ECOG')

# Apply to simulated data
sim_surv = os_sims_emax %>% 
  filter(.draw <= 500) %>%   # For speed in the class
  mutate(time=.value, event=rep(1,n())) %>% 
  nest(data = -.draw) %>% 
  mutate(km_est = map(data, ~vpc_stat_km(., pred_times=sort(unique(obs_surv$pred_times)), group='ECOG'))) %>% 
  select(-data) %>% 
  unnest(cols=km_est)

sim_surv %>% group_by(pred_times, group) %>% 
  summarise(med=mean(preds), 
            lcl = quantile(preds,probs = 0.05),
            ucl = quantile(preds, probs=0.95)) %>% 
  ggplot(aes(x=pred_times)) +
  geom_step(aes(y=med), color='red') +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.2) +
  geom_step(data=obs_surv,aes(y=preds)) +
  facet_wrap(~group) +
  ylim(0,1) + labs(x='Time (years)', y='Surviving fraction')
```

Survival VPC by RTS  quartile status 

```{r, echo=TRUE, out.height='50%'}
# Apply to observed data
obs_surv = vpc_stat_km(dos84 %>% mutate(time=TIME/365, event=DV), group='rts84_f')

# Apply to simulated data
sim_surv = os_sims_emax %>% 
  filter(.draw <= 500) %>%   # For speed in the class
  mutate(time=.value, event=rep(1,n())) %>% 
  nest(data = -.draw) %>% 
  mutate(km_est = map(data, ~vpc_stat_km(., pred_times=sort(unique(obs_surv$pred_times)), group='rts84_f'))) %>% 
  select(-data) %>% 
  unnest(cols=km_est)

sim_surv %>% group_by(pred_times, group) %>% 
  summarise(med=mean(preds), 
            lcl = quantile(preds,probs = 0.05),
            ucl = quantile(preds, probs=0.95)) %>% 
  ggplot(aes(x=pred_times)) +
  geom_step(aes(y=med), color='red') +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.2) +
  geom_step(data=obs_surv,aes(y=preds)) +
  facet_wrap(~group) +
  ylim(0,1) + labs(x='Time (years)', y='Surviving fraction')
```