---
title: "Introduction to programming with Stan"
output:
  slidy_presentation:
#    theme: cerulean
    fig_width: 9
    fig_height: 5
    font_adjustment: 3
    transition: none
    css: slidystyles.css
    footer: Metrum Research Group &copy 2021
    mathjax: local
    self_contained: false
    df_print: kable
    variant: markdown+fancy_lists
bibliography: Day3-references.bib
csl: statistical-science.csl
#nocite: | 
#    @2795, @2192, @3684
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE, echo=FALSE)

library(tidyverse)
library(stringr)
library(haven)
library(survival)
library(survminer)
library(survMisc)
library(texreg)
library(tidybayes)
library(bayesplot)
library(muhaz)
library(rstan)

set.seed(314159)
theme_set(theme_bw())
bayesplot::color_scheme_set("viridis")
``` 


```{r}
# OS data
d <- read_csv('../data/source/DDmodel0218_Simulated_OS.csv', na = c('.','-99'))

# Add week 12 (Day 84) predicted tumor size
d84 <- d %>% 
  filter(TIME <= 84) %>% 
  group_by(ID) %>% 
  mutate(rate = KG/1000 - KD0/1000*AUC0 - KD1/100*AUC1,
         prevTIME = lag(TIME,default = 0),
         change = exp(rate * (TIME-prevTIME)),
         ipred = IBASE * 1000 * cumprod(change)
         ) %>% 
  arrange(ID,TIME) %>% 
  slice(n()) %>% 
  mutate(ipred84 = ipred * exp(rate * (84-TIME)),
         rts84 = ipred84 / ( IBASE * 1000 ) )

dos <- d %>% 
  filter(TIME>0) %>% 
  group_by(ID) %>% 
  mutate(meanGem = mean(AUC1),
         Group = if_else(meanGem > 0, "Cb+G", "Cb")) %>% 
  ungroup() %>% 
  filter(CMT==2, EVID==0) %>% 
  left_join(d84 %>% select(ID, ipred84, rts84)) %>% 
  mutate(rts84_f = paste0("Q", ntile(rts84, n = 4)))

dos84 <- dos %>% 
  filter(TIME>84) %>% 
  mutate(TIME = TIME-84)


```

# Key learning objectives for today

* Introduction to Stan

* Extending TTE models to include
    * Non-linear effects
    * Repeated TTE

* Coming next week
  - Variables which vary continuously with time


# Stan is a probabilistic programming language

Links to the

* [Stan webpage](https://mc-stan.org/)
* [Stan reference manual](https://mc-stan.org/docs/2_21/reference-manual/)
* [Stan user's guide](https://mc-stan.org/docs/2_21/stan-users-guide/index.html)
* [Stan function reference](https://mc-stan.org/docs/2_21/functions-reference/index.html)

# HMC and NUTS



# Stan programs

A Stan program is organized into <red>a sequence of named blocks</red>. 

In each block:

* variable declarations
* followed by programming statements

All block types are optional but they must enter the model in a specific order.

Every variable used in a Stan program must have a declared data type

Variables declared in each block have scope over all subsequent statements (including statements in other blocks)

* One excepton: variables declared in model block are local to the model block

# Code blocks

::::{.columns}

:::{.column}

The following are valid code blocks and must appear in this order:

* functions
* <red>data</red>
* transformed data
* <red>parameters </red>
* transformed parameters
* <red> model </red>
* generated quantities

:::

:::{.column}

```
<block type>{
... declarations ...
... statements ...
}
```

:::
::::

# Data block

:::{.columns}

::::{.column}

* Declaration of variables that are read in as data
* Allows only declarations, no programming statements

```
data {
  int<lower=1> N;  // total number of observations
  int<lower=0, upper=1> Y[N];  // binary response variable
  vector[N] x;  // predictor
  int prior_only;  // should the likelihood be ignored?
}
```

::::

::::{.column}

Variable types:

* `int` for integer values
* `real` for continuous values
* `vector` for column vectors (real)
* `row_vector` for row vectors (real)
* `matrix` for matrices (real)

Any type can be made into an array type by declaring array arguments. For example,
```
real x[10];
matrix[3, 3] m[6, 7];
```

* `x` is a one-dimensional array of size $10$ containing real values
* `m` is a two-dimensional array of size  $6 \times 7$ containing values that are $3 \times 3$ matrices

NB: Array declaration is clearer in Version 2.26.

::::
:::

# Transformed parameters block

* For declaring and defining variables that that are functions of the data and that do not need to be changed
* Variables are defined in programming statements statements

```
transformed data {
  int Kc = K - 1;
  matrix[N, Kc] Xc;  // centered version of X without an intercept
  vector[Kc] means_X;  // column means of X before centering
  for (i in 2:K) {
    means_X[i - 1] = mean(X[, i]);
    Xc[, i - 1] = X[, i] - means_X[i - 1];
  }
}
```

# Parameters block

* The variables defined in this block are the parameters being sampled by Stan
* HMC and NUTS are implemented over a multivariate probability density that has support on all of $\mathbb{R}^n$

* Hamiltonian Monte Carlo requires the gradient of the (unnormalized) log probability function with respect to the unconstrained parameters 
    * the unconstrained version of the parameters is what’s sampled or optimized.

```
parameters {
  real alpha;  // intercept
  real beta ;  // slope
}
```


# Transformed parameters block

* Similar to the `transformed parameters` block but for parameters
* Variables defined in terms of data and parameters 



# Model block

* Where the log probability function is defined. 
* All variables defined here are local to the model block
    * May not be defined with constraints 

```
model {
  // log-likelihood contributions
  if (prior_only==0) {
    Y ~ bernoulli_logit(alpha + beta*x);
  }
  // prior didstributions
  beta ~ normal(0,3);
  alpha ~ normal(0,3);
}
```

# Generated quantities block

* Derived quantities based on parameters, data, and (optionally) random number generation
* Applications of posterior inference that can be coded in the generated quantities block include:
    * forward sampling to generate simulated data for model testing,
    * generating predictions for new data,
    * calculating posterior event probabilities
    * calculating posterior expectations
    * calculating log likelihoods, deviances, etc. for model comparison.

```
generated quantities {
  vector[N] Ysim;  // simulated response variable

  for (n in 1:N) {
    Ysim[n] = bernoulli_logit_rng(alpha + beta*x[n]);
  }
  
}
```


# Functions block

* A block to declare user-defined functions
* Function definitions and declarations may appear in any order, subject to the condition that a function must be declared before it is used. 

For example,

```
real cv(real mu, real sigma) { return(mu/sigma) };
```

A function that returns a `real` with two arguments of types `real` and `real` named `mu` and `sigma`.


# Example logistic regression model

* As an example, let's model the probability of ECOG status=1 as a function of observed baseline tumor size

$$
\text{logit}\left(P(\text{ECOG}=1 | x)\right) = \alpha + \beta  x
$$

* Stan code for this model is in `../model/stan/logistic_regression_example.stan`

# Running the model using rstan

:::{.column}
::::{.column}

* Define data as a list
* Provide initial values for Markov chains (optional)
* Call Stan

::::

::::{.column}

```{r, echo=TRUE, warning=FALSE, message=FALSE}
stan_data <- list(N = nrow(dos),
                  Y = dos$ECOG,
                  x = dos$SLD0/10,
                  prior_only = 0)

stan_inits <- function() {
  list(
    alpha = rnorm(1, mean=0, sd=2),
    beta = rnorm(1, mean=0, sd = 1)
  )
}

fit_lr <- stan(file = '../model/stan/logistic_regression_example.stan',
               data = stan_data,
               chains = 4,iter = 2000, warmup = 1000, 
               init = stan_inits, 
               cores = 2)
```

::::
:::

# Stan output

:::{.columns}

::::{.column, width='60%'}

```{r, echo=TRUE}
print(fit_lr, pars = c('alpha','beta'))
```
::::

::::{.column, width='40%'}

```{r, out.width='95%'}
plot(fit_lr, pars=c('alpha','beta'))
```

::::
:::

# Diagnostic plots

:::{.columns}

::::{.column}
```{r, echo=TRUE}
samples <- spread_draws(fit_lr, alpha,beta) %>% rename(Chain = .chain)
mcmc_trace(samples, pars = c('alpha','beta'))
```
::::

::::{.column}
```{r, echo=TRUE}
mcmc_scatter(samples, pars=c('alpha','beta'))
```

::::
:::


# Workbook Stan01

Expanding the logistic regression model to include a power term

$$
\text{logit}\left(P(\text{ECOG}=1 | x)\right) = \alpha + \beta  x^\gamma
$$

# Time to event model with a simple hazard model

Let's re-fit the overall survival model, using a weibull baseline hazard and a proportional effect of covariates.  

* Note this is a different parameterization than used in `brms`

$$
\begin{align*}
h_i(t) &= h_0(t) \times \exp(\beta \times \text{RTS}_i)  \\
h_0(t) &= \lambda_0 \times \alpha \ \times t^{\alpha-1}\\
\\
H_i(t) &= \lambda_i t^\alpha \\
S_i(t) &= \exp(-\lambda_i t^\alpha)
\end{align*}
$$

Under the standard assumption that censoring times are independent of event times, the individual contributions to the log-likelihood function are

$$
\ell_i(\theta) = \begin{cases}
\log h_i(T^*_i) - H_i(T^*_i) & \text{for } \delta_i = 1 \\
-H(T^*_i) & \text{for } \delta_i = 0
\end{cases}
$$

# Model fitting

The code is in `../model/stan/weibull_example_v2.stan`.

The Rmarkdown file for running the example is in `weibull_example.Rmd`



# Integrating the hazard function

Stan has some built-in functions to do integration.

We'll start with the simple one-dimensional case before moving to systems of ODEs.

# Function block to define the integrand

For comparison, we'll have Stan integrate the Weibull hazard
  * Not something we'd do in practice
  * Simply for comparison sake :)
  
Integrand functions need this signature, even if you don't use all of the arguments.
```
functions {
  
  real weibull_hazard(real x,        // Function argument
                    real xc,         // Complement of function argument
                                     //  on the domain (defined later)
                    real[] theta,    // parameters
                    real[] x_r,      // data (real)
                    int[] x_i) {     // data (integers)
                      
                      real sigma = theta[1];
                      real alpha = theta[2];
                      
                      return( (alpha/sigma) * pow(x/sigma, alpha-1) );
                      
                    }
}
```

# Use of integrate_1d function

Definite and indefinite one dimensional integrals can be performed in Stan using the [`integrate_1d`](https://mc-stan.org/docs/2_19/functions-reference/functions-1d-integrator.html) function.

```
for (n in 1:N) {

  // Evaluate hazard function - pass arguments to function
  hazard[n] = weibull_hazard(Y[n], 0.0, {mu[n],shape}, {0.0}, {0}  );

  // Integrate hazard function using integrate_1d
  // arguments: function, lower limit, upper limit, theta, x_r, x_i, relative tolerance
  cumulative_hazard[n] = integrate_1d(weibull_hazard, 1E-06, Y[n], {mu[n],shape}, {0.0}, {0}, 1E-08 ) ;

}
```

# Let's use this to fit a model and compare results

An example using the Weibull hazard is in `../model/stan/weibull_example_integrated_hazard.stan`

The Rmarkdown file for running the example is in `weibull_integrated_hazard_example.Rmd`


# Example

```{r, eval=FALSE}

stan_data <- list(N = nrow(dos84),
                  Y = dos84$TIME,
                  cens = 1 - dos84$DV,
                  K = 2,
                  X = model.matrix(~rts84, data=dos84),
                  prior_only = 0
                  )

stan_inits <- function() {
  list(
  # Give initial values that avoid numerical issues at t=0
    shape = runif(1, 1, 3)
  )
}

fit = stan(file='../model/stan/weibull_integrated_hazard.stan', 
           data = stan_data,
           init = stan_inits,
           chains=3, iter=1000)
```






# Numeric model comparison

* AIC and friends
* C-index

# Workbook 5: Parametric models



# Stan

* Every variable used in a Stan program must have a declared data type.

