---
title: "Workbook for Introduction to TTE modeling"
author: "Weibull TTE models in Stan"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(survival)
library(survminer)
library(texreg)
library(mgcv)
library(muhaz)
library(rstan)
library(bayesplot)
library(tidybayes)

theme_set(theme_bw())
bayesplot::color_scheme_set("viridis")

```


```{r}
# OS data
d <- read_csv('../data/source/DDmodel0218_Simulated_OS.csv', na = c('.','-99'))

# Add week 12 (Day 84) predicted tumor size
d84 <- d %>% 
  filter(TIME <= 84) %>% 
  group_by(ID) %>% 
  mutate(rate = KG/1000 - KD0/1000*AUC0 - KD1/100*AUC1,
         prevTIME = lag(TIME,default = 0),
         change = exp(rate * (TIME-prevTIME)),
         ipred = IBASE * 1000 * cumprod(change)
         ) %>% 
  arrange(ID,TIME) %>% 
  slice(n()) %>% 
  mutate(ipred84 = ipred * exp(rate * (84-TIME)),
         rts84 = ipred84 / ( IBASE * 1000 ) )

dos <- d %>% 
  filter(TIME>0) %>% 
  group_by(ID) %>% 
  mutate(meanGem = mean(AUC1),
         Group = if_else(meanGem > 0, "Cb+G", "Cb")) %>% 
  ungroup() %>% 
  filter(CMT==2, EVID==0) %>% 
  left_join(d84 %>% select(ID, ipred84, rts84)) %>% 
  mutate(rts84_f = paste0("Q", ntile(rts84, n = 4)))

dos84 <- dos %>% 
  filter(TIME>84) %>% 
  mutate(TIME = TIME-84)


```


# Working with traditional pharmacometric parameterization of Weibull model


$$
\begin{align*}
h_i(t) &= h_0(t) \times \exp(\beta \times \text{RTS}_i)  \\
h_0(t) &= \lambda_0 \times \alpha \ \times t^{\alpha-1}\\
\\
H_i(t) &= \lambda_i t^\alpha \\
S_i(t) &= \exp(-\lambda_i t^\alpha)
\end{align*}
$$
Under this parameterization, 

* median OS = $\left( \frac{\log 2}{\lambda_0 \exp(\beta \times RTS)} \right)^{1/\alpha}$
* mean OS = $\left\{ \lambda_0 \exp(\beta \times RTS) \right\}^{-1/\alpha} ~~\Gamma(1 + 1/\alpha)$


Set-up the data

```{r}

tte_data <- list(
  N = nrow(dos84),
  Y = dos84$TIME/365,  # Use years instead of days
  cens = 1-dos84$DV,
  K = 1,
  X = model.matrix(~-1 + rts84, data=dos84),
  prior_only = 0
)
```

Set up a function to generate the initial values:

```{r}
weibull_inits <- function() {
  list(
#    b = rnorm(tte_data$K,0,2),
    lambda0 = rlnorm(1,0,5),
    shape = rlnorm(1,0,1)
    )
}

```

Run the model


```{r, results='hide'}
fit_weibull_pmx <- stan(file = '../model/stan/weibull_example_v2.stan',
                        data = tte_data, 
                        init = weibull_inits,
                        chains = 4,
                        cores = 2,
                        seed = 76341
                        )
```


Look at the output

```{r}
print(fit_weibull_pmx)
```

Try again with the parameters of interest

```{r}
print(fit_weibull_pmx, pars=c('lambda0','shape','b'))
```

To get a more precise summaery, use summary:


```{r}
summary(fit_weibull_pmx, pars=c('lambda0','shape','b'))$summary
```


How does the effect of `rts84` compare to the estimate from the `brms` model?  
Why do you think they differ?



Now let's pull out the simulated event times from our model (simulated in the `generated quantities` block):

```{r}
os_sims <- gather_draws(fit_weibull_pmx, Ysim[ID]) %>% 
  left_join(dos84)
head(os_sims)
```

Let's make a quick plot to assess model fit, ignoring censoring for now:

```{r}

vpc_stat_km <- function(.data, group='rts84_f', pred_times=NULL) {
  .data$group <- .data[[group]]
  
  fit <- survfit(Surv(time,event)~group, data=.data)
  if (is.null(pred_times)) {
    pred_times <- c(0,sort(unique(fit$time)))
  }
  preds = summary(fit, times=pred_times)
  
  data.frame(pred_times=preds$time, preds = preds$surv, group=preds$strata)
}

# Apply to observed data
obs_surv = vpc_stat_km(dos84 %>% mutate(time=TIME/365, event=DV))

# Apply to simulated data
sim_surv = os_sims %>% 
  filter(.draw <= 500) %>%   # For speed in the class
  mutate(time=.value, event=rep(1,n())) %>% 
  nest(data = -.draw) %>% 
  mutate(km_est = map(data, ~vpc_stat_km(., pred_times=sort(unique(obs_surv$pred_times))))) %>% 
  select(-data) %>% 
  unnest(cols=km_est)
```

# Plot survival function VPC stratified by RTS84 quartile

```{r, echo=TRUE, out.height='50%'}
sim_surv %>% group_by(pred_times, group) %>% 
  summarise(med=mean(preds), 
            lcl = quantile(preds,probs = 0.05),
            ucl = quantile(preds, probs=0.95)) %>% 
  ggplot(aes(x=pred_times)) +
  geom_step(aes(y=med), color='red') +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.2) +
  geom_step(data=obs_surv,aes(y=preds)) +
  facet_wrap(~group) +
  ylim(0,1) + labs(x='Time (years)', y='Surviving fraction')
```

# Plot survival function VPC stratified by ECOG status

```{r, echo=TRUE, out.height='50%'}
# Apply to observed data
obs_surv = vpc_stat_km(dos84 %>% mutate(time=TIME/365, event=DV), group='ECOG')

# Apply to simulated data
sim_surv = os_sims %>% 
  filter(.draw <= 500) %>%   # For speed in the class
  mutate(time=.value, event=rep(1,n())) %>% 
  nest(data = -.draw) %>% 
  mutate(km_est = map(data, ~vpc_stat_km(., pred_times=sort(unique(obs_surv$pred_times)), group='ECOG'))) %>% 
  select(-data) %>% 
  unnest(cols=km_est)

sim_surv %>% group_by(pred_times, group) %>% 
  summarise(med=mean(preds), 
            lcl = quantile(preds,probs = 0.05),
            ucl = quantile(preds, probs=0.95)) %>% 
  ggplot(aes(x=pred_times)) +
  geom_step(aes(y=med), color='red') +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.2) +
  geom_step(data=obs_surv,aes(y=preds)) +
  facet_wrap(~group) +
  ylim(0,1) + labs(x='Time (years)', y='Surviving fraction')
```


** Exercise 1:**

1. Run the model adding an effect of ECOG status.
  a. Do you need to modify the model or just the data?
  b. Does this improve the model fit with respect to baseline ECOG status?
  
  
** Exercise 2:**

2. Now expand the model to include a non-linear effect of RTS84
  a. Try an Emax relationship as shown below.  Does this model improve the fit at all?
  I suggest you try to code this on your own, but if you get stuck, the code below walks through an answer.
  
$$
  h_i(t) = h_0(t) \times \exp \left( \frac{\beta_1 \times \text{RTS}_i}{E50 + \text{RTS}_i} + \beta_2 \text{ECOG} \right)
$$

```{r}
tte_emax_data <- list(
  N = nrow(dos84),
  Y = dos84$TIME/365,  # Use years instead of days
  cens = 1-dos84$DV,
  K = 1,
  X = model.matrix(~-1 + ECOG, data=dos84),
  RTS = dos84$rts84,
  prior_only = 0
)
```



```{r, results='hide'}
fit_weibull_emax <- stan(file = '../model/stan/weibull_example_emax.stan',
                        data = tte_emax_data, 
                        chains = 4,
                        cores = 2,
                        seed = 76341
                        )
```

```{r}
samples_emax <- spread_draws(fit_weibull_emax, emax,e50,b[i],lambda0,shape) %>% 
  rename(Chain=.chain)
```

```{r}
mcmc_trace(samples_emax, pars=c('emax','e50','b','lambda0','shape'))
```

```{r}
mcmc_pairs(samples_emax, pars=c('emax','e50','b','lambda0','shape'))
```

Compare models using loo


```{r}
loo_pmx <- loo::loo(extract(fit_weibull_pmx, pars='log_lik')$log_lik)
loo_emax <- loo::loo(extract(fit_weibull_emax, pars='log_lik')$log_lik)

loo_compare(loo_pmx, loo_emax)

```
Plot estimated relationship with RTS

```{r}
RTS_grid <- seq(min(dos84$rts84), max(dos84$rts84), length=100)

samples_emax %>% crossing(rts = RTS_grid) %>% 
  mutate(log_hr = emax*rts / (e50+rts)) %>% 
  group_by(rts) %>% 
  summarise(est = median(log_hr), 
            lcl=quantile(log_hr,probs=0.05),
            ucl = quantile(log_hr, probs = 0.95)) %>% 
  ggplot(aes(x=rts, y=est)) +
  geom_line() +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.2) +
  labs(x='Relative change in tumor size to week 12', y='log hazard ratio') 
```

Simple VPCs without censoring

```{r}
os_sims_emax <- gather_draws(fit_weibull_emax, Ysim[ID]) %>% 
  left_join(dos84)
head(os_sims_emax)
```

Survival VPC by ECOG status 

```{r, echo=TRUE, out.height='50%'}
# Apply to observed data
obs_surv = vpc_stat_km(dos84 %>% mutate(time=TIME/365, event=DV), group='ECOG')

# Apply to simulated data
sim_surv = os_sims_emax %>% 
  filter(.draw <= 500) %>%   # For speed in the class
  mutate(time=.value, event=rep(1,n())) %>% 
  nest(data = -.draw) %>% 
  mutate(km_est = map(data, ~vpc_stat_km(., pred_times=sort(unique(obs_surv$pred_times)), group='ECOG'))) %>% 
  select(-data) %>% 
  unnest(cols=km_est)

sim_surv %>% group_by(pred_times, group) %>% 
  summarise(med=mean(preds), 
            lcl = quantile(preds,probs = 0.05),
            ucl = quantile(preds, probs=0.95)) %>% 
  ggplot(aes(x=pred_times)) +
  geom_step(aes(y=med), color='red') +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.2) +
  geom_step(data=obs_surv,aes(y=preds)) +
  facet_wrap(~group) +
  ylim(0,1) + labs(x='Time (years)', y='Surviving fraction')
```

Survival VPC by RTS  quartile status 

```{r, echo=TRUE, out.height='50%'}
# Apply to observed data
obs_surv = vpc_stat_km(dos84 %>% mutate(time=TIME/365, event=DV), group='rts84_f')

# Apply to simulated data
sim_surv = os_sims_emax %>% 
  filter(.draw <= 500) %>%   # For speed in the class
  mutate(time=.value, event=rep(1,n())) %>% 
  nest(data = -.draw) %>% 
  mutate(km_est = map(data, ~vpc_stat_km(., pred_times=sort(unique(obs_surv$pred_times)), group='rts84_f'))) %>% 
  select(-data) %>% 
  unnest(cols=km_est)

sim_surv %>% group_by(pred_times, group) %>% 
  summarise(med=mean(preds), 
            lcl = quantile(preds,probs = 0.05),
            ucl = quantile(preds, probs=0.95)) %>% 
  ggplot(aes(x=pred_times)) +
  geom_step(aes(y=med), color='red') +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.2) +
  geom_step(data=obs_surv,aes(y=preds)) +
  facet_wrap(~group) +
  ylim(0,1) + labs(x='Time (years)', y='Surviving fraction')
```