---
title: "Workbook for Introduction to modeling with Stan"
author: "Logistic regression models"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  # html_document:
  #   css: docs/src/styles/styles.css
  #   number_sections: true
  #   theme: united
  #   toc: true
  #   toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(mgcv)
library(rstan)
library(tidybayes)
library(bayesplot)

theme_set(theme_bw())
bayesplot::color_scheme_set("viridis")
```


We'll continue to use the simulated data from Zecchin et al., as posted on the DDMoRe repository.

The objective of this analysis is to explore the relationship between ECOG status (0 vs 1) and baseline tumor size.  We will fit two models, compare them using LOO-IC and VPCs.

We'll start by reading the data

```{r}
# OS data
d <- read_csv('../data/source/DDmodel0218_Simulated_OS.csv', na = c('.','-99'))

# Add week 12 (Day 84) predicted tumor size
d84 <- d %>% 
  filter(TIME <= 84) %>% 
  group_by(ID) %>% 
  mutate(rate = KG/1000 - KD0/1000*AUC0 - KD1/100*AUC1,
         prevTIME = lag(TIME,default = 0),
         change = exp(rate * (TIME-prevTIME)),
         ipred = IBASE * 1000 * cumprod(change)
         ) %>% 
  arrange(ID,TIME) %>% 
  slice(n()) %>% 
  mutate(ipred84 = ipred * exp(rate * (84-TIME)),
         rts84 = ipred84 / ( IBASE * 1000 ) )

dos <- d %>% 
  filter(TIME>0) %>% 
  group_by(ID) %>% 
  mutate(meanGem = mean(AUC1),
         Group = if_else(meanGem > 0, "Cb+G", "Cb")) %>% 
  ungroup() %>% 
  filter(CMT==2, EVID==0) %>% 
  left_join(d84 %>% select(ID, ipred84, rts84)) %>% 
  mutate(rts84_f = paste0("Q", ntile(rts84, n = 4)))

dos84 <- dos %>% 
  filter(TIME>84) %>% 
  mutate(TIME = TIME-84)


```

Next we'll fit the model from the slides, namely

$$
\text{logit}\left(P(\text{ECOG}=1 | x)\right) = \alpha + \beta  x
$$

Open the file `../model/stan/logistic_regression_example.stan`.  Look over the four code blocks.  Do you understand what each of these is doing?  If not, ask :)

We'll start by doing a prior-predictive check to see if our prior distributions make sense.  To do this, we'll set `prior_only=1` in out data.  Based on your review of the code, what do you expect this to do?

Use the code below to put together the data for the model, set up the initial values and run the model.  

```{r, echo=TRUE, warning=FALSE, message=FALSE}
stan_data <- list(N = nrow(dos),
                  Y = dos$ECOG,
                  x = dos$SLD0/10,
                  prior_only = 1)

stan_inits <- function() {
  list(
    alpha = rnorm(1, mean=0, sd=2),
    beta = rnorm(1, mean=0, sd = 1)
  )
}

fit_lr <- stan(file = '../model/stan/logistic_regression_example.stan',
               data = stan_data,
               chains = 4,iter = 2000, warmup = 1000, 
               init = stan_inits, 
               cores = 2,
               seed = 31425)
```


To assess the prior predictive relationship, we've just simulated from the prior distribution.  Now we'll plot some of the sampled relationships between baseline tumor size and the probability of ECOG=1.

```{r}
prior_samples = spread_draws(fit_lr, alpha, beta)
head(prior_samples)
```
 Let's plot 100 curves
 
```{r}
prior_samples %>% 
  sample_n(size=200) %>% 
  crossing(SLD0 = seq(min(dos$SLD0), max(dos$SLD0), length=100)) %>% 
  ggplot(aes(x=SLD0, y=plogis(alpha + beta*SLD0), group=.draw)) + 
  geom_line(alpha=0.5) +
  labs(x='Baseline tumor size', y='P(ECOG=1)')
```
 Does this seem reasonable?  If not, how might you modify the prior distributions to get something that seems more reasonable to you?  Try modifying the priors and re-running the prior predictive check.
 
Once you're happy with your priors, we'll fit the model to the data (to get samples from the posterior distribution).  Note the change of `prior_only` from 1 to 0.


```{r, echo=TRUE, warning=FALSE, message=FALSE}
stan_data <- list(N = nrow(dos),
                  Y = dos$ECOG,
                  x = dos$SLD0/10,
                  prior_only = 0)

fit_lr <- stan(file = '../model/stan/logistic_regression_example.stan',
               data = stan_data,
               chains = 4,iter = 2000, warmup = 1000, 
               init = stan_inits, 
               cores = 2, seed = 31425)
```

```{r}
posterior_samples = spread_draws(fit_lr, alpha, beta) %>% 
  rename(Chain = .chain)  # The bayesplot functions look for a variable `Chain`
head(posterior_samples)
```
First, we'll check for convergence of the Markov chains by looking at the Rhat values and trace plots

```{r}
print(fit_lr, pars=c('alpha','beta'))
```

```{r}
mcmc_trace(posterior_samples, pars=c('alpha','beta'))
```
Based on these diagnostics, do you think the sampler has converged?

Let's look at a VPC as a function of baseline tumor size.  First, we'll define our summary function.

```{r}
summary_function <- function(.data, .x_name, .y_name='value') {
  .data <- .data %>% ungroup() %>% rename('xvar' = .x_name, 'yvar'=.y_name)
  x_grid <- with(.data, seq(from = min(xvar),
                           to = quantile(xvar,probs = 0.95), 
                           length = 100))
  fit <- gam(yvar ~ s(xvar), family=binomial(link='logit'), data=.data)
  predictions <- predict(fit, newdata = data.frame(xvar=x_grid), type='response')
  return( data.frame(xvar=x_grid, prediction = predictions))
}
```

Next, we'll compute the summary statistics for the observed data

```{r}
obs_summary <- summary_function(dos, .x_name = 'SLD0', .y_name='ECOG') %>% 
  mutate(type='Observed')

head(obs_summary)
```

Next, we'll extract the posterior predictive samples and calculate the smooth relationship for each simulated dataset

```{r}
posterior_predictive_samples = gather_draws(fit_lr, Ysim[ID]) %>% 
  rename(Chain = .chain)

head(posterior_predictive_samples)
```


```{r}
sim_summary <- posterior_predictive_samples %>% 
  # Join original data
  left_join(dos %>% select(ID,SLD0)) %>% 
  # Nest everying except the simulation name
  nest(cols=-.draw) %>% 
  # Use 200 sims for demonstration
  sample_n(size=200) %>% 
  # Compute summary stats for each simulated dataset
  mutate(predictions = map(cols, ~summary_function(.x, 
                                                   .x_name='SLD0',
                                                   .y_name='.value'))) %>% 
  select(.draw,predictions) %>% 
  unnest(cols=predictions) %>% 
  # Summarise across simulated data sets
  group_by(xvar) %>% 
  summarise(qlo = quantile(prediction, probs = 0.05),
            qhi = quantile(prediction, probs = 0.95),
            prediction=median(prediction)
            ) %>% 
  mutate(type = 'Simulated')
```

# Plot VPC

```{r, 'cont-vpc'}
sim_summary %>% bind_rows(obs_summary) %>% 
  ggplot(aes(x=xvar, y=prediction)) +
  geom_line(aes(col=type, group=type)) +
  geom_ribbon(aes(ymin=qlo, ymax=qhi, fill=type), alpha=0.2) +
  labs(x='Baseline SLD', y='P(ECOG=1)')

```

Let's plot 100 curves from the posterior
 
```{r}
posterior_samples %>% 
  sample_n(size=200) %>% 
  crossing(SLD0 = seq(min(dos$SLD0), max(dos$SLD0), length=100)) %>% 
  ggplot(aes(x=SLD0, y=plogis(alpha + beta*SLD0/10), group=.draw)) + 
  geom_line(alpha=0.5) +
  labs(x='Baseline tumor size', y='P(ECOG=1)')
```

Let's plot the estimated relationship and pointwise 90% credible interval

```{r}
posterior_samples %>% 
  sample_n(size=500) %>% 
  crossing(SLD0 = seq(min(dos$SLD0), max(dos$SLD0), length=100)) %>% 
  # NB: We passed SLD0 in cm into the Stan model...
  mutate(phat = plogis(alpha + beta*SLD0/10)) %>% 
  group_by(SLD0) %>% 
  mutate(est = median(phat), lcl=quantile(phat, probs = 0.05), ucl=quantile(phat, probs=0.95)) %>% 
  ggplot(aes(x=SLD0, y=est)) + 
  geom_line() +
  geom_ribbon(aes(ymin=lcl, ymax=ucl), fill='red', alpha=0.30) +
  ylim(0,1) +
  labs(x='Baseline tumor size', y='P(ECOG=1)')
```

*** Exercise: ***

1. Inspect the Stan model file `../model/stan/logistic_regression_power.stan`

2. See how the code was modified to use a power model for SLD0 i.e., $\text{logit}\left(P(\text{ECOG}=1 | x)\right) = \alpha + \beta  x^\gamma$.  Focus on the parameters, model, and generated quantities blocks.

3. Run the model.  Did you get some post-warmup divergent transitions?  If so, follow the link provided by the Stan output (http://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup) to read more.
  a. Look at the pairs plot for alpha, beta and the power parameter.  What do you see?  Try increasing the `adapt_delta` option to 0.95.  Does that fix the problem?

```{r}
fit_lr_power <- stan(file = '../model/stan/logistic_regression_power.stan',
               data = stan_data,
               chains = 4,iter = 2000, warmup = 1000, 
               cores = 2,  seed=31425)
```

  b. Try a model which forces $\beta > 0$, does that fix the problem?  Do you think this is a reasonable solution, or would you try a different functional form?

4. Check convergence and model fit using a VPC.  For the VPC, you can use the code above with no changes.

5. Compare models using LOO-IC.  Which model do you think will provide better predictions for future patients?

```{r}
loo_base <- loo(fit_lr)
loo_power <- loo(fit_lr_power)

print(loo_base)
print(loo_power)

loo::loo_compare(loo_base, loo_power)
```
