---
title: "Workbook for Introduction to TTE modeling"
author: "Cox regression"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  # html_document:
  #   css: docs/src/styles/styles.css
  #   number_sections: true
  #   theme: united
  #   toc: true
  #   toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(survival)
library(survminer)
library(texreg)
library(mgcv)
library(flexsurv)
library(muhaz)
library(Hmisc)

theme_set(theme_bw())
```


```{r, "TRY toydata", purl=TRUE, results = "hide"}
load('../data/aedat.RDS')

aedat <-
  aedat %>% 
  mutate(AETOXGR = factor(aedat$AETOXGR, 0:3, labels=c("None","Mild","Moderate","Severe")),
         ae_any = AETOXGR != 'None') %>% 
  group_by(USUBJID) %>%
  # End of study for patients without a severe event
  mutate(TTE_SEVERE = case_when(
    STUDYID=="PROTA" ~ 2,
    STUDYID=="PROTB" ~ 6
  ),
  # Time of severe event for those that had one
  TTE_SEVERE = ifelse(AETOXGR=="Severe", TTE, TTE_SEVERE)
  )

# Both for EDA and for model-checking, it's generally helpful to have quartiles of exposure:
dat_use <-
  aedat %>% arrange(USUBJID, TTE_SEVERE) %>% slice(1) %>%
  group_by(PBO) %>%
  mutate(Quartile = ifelse(PBO == "PBO", "PBO",
                           paste0("Q", ntile(CAVGSS, n = 4)))) %>% 
  ungroup() %>% 
  mutate(rowid = 1:n())

```


# Workbook 4:  Cox regression

We'll start by fitting a model to time to a severe event as a function of CAVGSS

```{r}
cox_model01 <- coxph(Surv(TTE_SEVERE, AE01) ~ CAVGSS, data=dat_use)
```

```{r}
print(cox_model01)
```


Let's look at the deviance residuals

* Versus body weight and patient type, to see if there is a relationship with covariates not in our model
* Versus CAVGSS to see if we've modeled the relationship reasonably well

```{r}
dat_use <- dat_use %>%
  mutate(resid01 = residuals(cox_model01, type='deviance'))
```

```{r}
dat_use %>% 
  ggplot(aes(x=BWT, y=resid01)) +
  geom_point() +
  geom_smooth() +
  labs(x='Body weight', y='Deviance residual')
```

```{r}
dat_use %>% 
  ggplot(aes(x=PTTYPE, y=resid01)) +
  geom_violin() +
  geom_point() +
  labs(x='Patient type', y='Deviance residual')
```

```{r}
dat_use %>% 
  ggplot(aes(x=CAVGSS, y=resid01)) +
  geom_point() +
  geom_smooth() +
  labs(x='Body weight', y='Exposure')
```

Do you see any evidence that we should modify our model based on these plots?  If so, what leads you to that decision and how might you change the model?

Let's make some predictions and overlay the model fits

```{r}
preds_mod01 <- survfit(cox_model01, newdata = dat_use) %>% survfit0()
```

```{r}
# Make a tall version of the data
survival_preds_fit1 <- preds_mod01$surv %>% 
  as.data.frame() %>% 
  mutate(time = preds_mod01$time) %>% 
  pivot_longer(cols=-time) %>% 
  # Add a row id column for merging with the covariate data
  mutate(rowid = as.numeric(name))

# Merge predictions and covariates for plotting
survival_preds_fit1 <- survival_preds_fit1 %>% 
  left_join(dat_use) %>% 
  # Because CAVGSS is continuous, we need to plot using categories
  # So, we'll use exposure quartiles
  group_by(time, Quartile) %>% 
  # Calculare expected value for S(t) across subjects in each group
  summarise(est=mean(value))

# Plot the observed data (the Kaplan-Meier estimate)
ggsurvplot(survfit(Surv(TTE_SEVERE, AE01) ~ Quartile, data=dat_use), 
           data=dat_use)$plot +
  # Overlay the predictions
  geom_step(data=survival_preds_fit1,
            aes(x=time, y=est), linetype='dashed') +
  facet_wrap(~Quartile
             )
```

Does this plot align with your expectation based on the residual plots?

**Exercise**

* Make similar plots for patient type and body weight.  Do these align with your expectation based on the residual plots?
   - Hint: you'll need to make a categorical variable from body weight.  The `ntile()` function is a handy tool for this.

```{r}
# Your code goes here
```


Now, let's fit three more models for comparison

* Surv(TTE_SEVERE,AE01) ~ CAVGSS + PTTYPE
* Surv(TTE_SEVERE, AE01) ~ Quartile
* Surv(TTE_SEVERE,AE01) ~ pspline(CAVGSS, df=4)

The last model will fit a non-parametric (smooth) function of CAVGSS using penalized splines with 4 degrees of freedom.

```{r}
cox_model02 <- update(cox_model01, . ~ . + PTTYPE)
cox_model03 <- coxph(Surv(TTE_SEVERE, AE01) ~ Quartile, data=dat_use)
cox_model04 <- coxph(Surv(TTE_SEVERE, AE01) ~ pspline(CAVGSS, df=4), data=dat_use)
```

To get an idea of how the effects look on the log-hazard scale, we can use the termplot function

```{r}
termplot(cox_model04,terms = 1)
```


We'll compare these models initially using concordance.

```{r}
concordance(cox_model01,cox_model02,cox_model03,cox_model04)
```


**Exercise**

1. Based on the c-index, is there one model that is clearly better than the others?  

2. For comparison, 
    a. Plot the deviannce residuals from model `cox_mod04` vs CAVGSS.  Do they look better than the residuals from model 01?
    b. Make the observed and predicted plot for model `cox_mod04`.  How do these compare to `cox_mod01`?

3. Based on all of this information, which of these models would you select going forward? Why?



