---
title: "Workbook for Introduction to TTE modeling"
author: "Kaplan-Meier estimates and plots"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  # html_document:
  #   css: docs/src/styles/styles.css
  #   number_sections: true
  #   theme: united
  #   toc: true
  #   toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(survival)
library(survminer)
library(texreg)
library(mgcv)
library(flexsurv)
library(muhaz)
library(Hmisc)

theme_set(theme_bw())
```

# Estimating the survival function 

We'll use the `survfit` function to estimate the Survival function.  

Using our example data from the last class, let's estimate the overall S(t), i.e., not stratifting by any covariates.

```{r, "TRY toydata", purl=TRUE, results = "hide"}
load('../data/aedat.RDS')

aedat <-
  aedat %>% 
  mutate(AETOXGR = factor(aedat$AETOXGR, 0:3, labels=c("None","Mild","Moderate","Severe")),
         ae_any = AETOXGR != 'None') %>% 
  group_by(USUBJID) %>%
  # End of study for patients without a severe event
  mutate(TTE_SEVERE = case_when(
    STUDYID=="PROTA" ~ 2,
    STUDYID=="PROTB" ~ 6
  ),
  # Time of severe event for those that had one
  TTE_SEVERE = ifelse(AETOXGR=="Severe", TTE, TTE_SEVERE)
  )

# Both for EDA and for model-checking, it's generally helpful to have quartiles of exposure:
dat_use <-
  aedat %>% arrange(USUBJID, TTE_SEVERE) %>% slice(1) %>%
  group_by(PBO) %>%
  mutate(Quartile = ifelse(PBO == "PBO", "PBO",
                           paste0("Q", ntile(CAVGSS, n = 4)))) %>% 
  ungroup() %>% 
  mutate(rowid = 1:n())

```


The key elements:

* `survfit`: the function used to obtain the K-M (or Flemming-Harrington) estimate of the survival function
  - stype=1 (the default)   gives the K-M estimate
  - stype=2 gives the Fleming-Harrington estimate

* `Surv` function for defining the outcome
  - For right-censored data, the first argument is the observed event time variable, and the second argument is the even indicator (1 or TRUE for an event; 0 or FALSE for censoring)
  
Obtaining the estimate and looking at some output:

```{r}
km_est = survfit(Surv(TTE,ae_any)~1, data = dat_use)

print(km_est)
summary(km_est)
```


Plotting the estimate using `ggsurvplot`:

```{r}
ggsurvplot(km_est, risk.table = TRUE)
```

Often it's helpful to add the number of subjects at risk to the bottom of the plot (a risk table).

**Exercise:**

1. Obtain the Kaplan-Meier estimates for time to any AE by exposure quartile.  Hint: the righ hand side of the formula will be "~ Quartile"

2. Plot the estimates using ggsurvplot.  Do you see any evidence of an exposure-response relationship?

3. Repeat steps 1 and 2 for time to a severe AE (event time variable is TTE_SEVERE, and the event indicator variable is AE01). Do you see any evidence of an exposure-response relationship?

3. Extra: Plot the K-M estimates by patient type, faceted by exposure quartile. Hint: Fit model `Surv(TTE,AY_any)~PTTYPE` and use the `facet.by` argument to `ggsurvplot`

```{r}
# Your code goes here
```

