---
title: "Workbook for Introduction to TTE modeling"
author: "Relationship between functions"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  # html_document:
  #   css: docs/src/styles/styles.css
  #   number_sections: true
  #   theme: united
  #   toc: true
  #   toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(survival)
library(survminer)
library(texreg)
library(mgcv)
library(flexsurv)
library(muhaz)
library(Hmisc)

theme_set(theme_bw())
```


# Plotting hazard and survival functions over time for some common distributions

To better understand the relationship between the hazard, cumulative hazard, survival and density functions, let's plot them for a three commonly used parametric distributions in TTE modeling: exponential, Weibull, and log-normal.

## Exponential distribution

The hazard is constant as a funtion of time: $h(t) = \lambda$.

From first principles:
 * h(t) = lambda
 * H(t) = lambda * t
 * S(t) = exp(-lambda * t)
 * f(t) = h(t) * S(t) = lambda * exp(-lambda * t)
 
```{r}
plot_data <- crossing(lambda = c(0.5,1,2), times = seq(0,5,length=100)) %>% 
  mutate(hazard = lambda,
         cumulative_hazard = lambda * times,
         survival = exp(-cumulative_hazard),
         density = hazard * survival)

plot_data %>% 
  pivot_longer(cols=hazard:density, names_to = 'type', values_to='value') %>% 
  ggplot(aes(x=times, y=value, col=as.factor(lambda), group=lambda)) +
  geom_line() +
  facet_wrap(~type, scales = "free_y")
```


## Weibull

The Weibull distribution has two parameters: lambda and gamma

The log hazard is linear in the log of time: 
$$h(t) = \gamma \lambda t^{\gamma-1} \iff \log h(t) = \log\gamma + \log \lambda + (\gamma - 1) \log t$$


From first principles:
 * h(t) = lambda * gamma * t^(gamma-1)
 * H(t) = lambda * t^gamma
 * S(t) = exp(-lambda * t^gamma)
 * f(t) = h(t) * S(t) = lambda * gamma * t^(gamma-1) * exp(-lambda * t^gamma)

```{r}
plot_data <- crossing(lambda = c(0.5,2), gamma = c(0.5, 1, 3), times = seq(0,5,length=100)) %>% 
  mutate(hazard = lambda * gamma * times^(gamma-1),
         cumulative_hazard = lambda * times^gamma,
         survival = exp(-cumulative_hazard),
         density = hazard * survival
)

plot_data %>% 
  pivot_longer(cols=hazard:density, names_to = 'type', values_to='value') %>% 
  mutate(combination = paste(lambda,gamma,sep='-')) %>% 
  ggplot(aes(x=times, y=value, col=as.factor(gamma), group=combination)) +
  geom_line() +
  facet_grid(type~lambda, scales = "free_y")
```

## Log-normal

The hazard isn't easily expressed in closed form.  How can you derive it from first principles?  Edit the code below to make the plots.

Hint: the log normal pdf and survival function are obtained using the `dlnorm` and `plnorm` functions, respectively.

Plot the log-normal hazard function for combinations of mean (1, 2) and starndard deviation (0.3,1) modifying the code below.



```{r}
plot_data <- crossing(mean = c(1,2), sd = c(0.3, 1), times = seq(0,5,length=100)) %>% 
  mutate(survival = "insert survival function",
         density = "insert density function",
         cumulative_hazard = "insert cumulative hazard function",
         hazard = "insert hazard function"
)

plot_data %>% 
  pivot_longer(cols=survival:hazard, names_to = 'type', values_to='value') %>% 
  mutate(combination = paste(mean,sd,sep='-')) %>% 
  ggplot(aes(x=times, y=value, col=as.factor(sd), group=combination)) +
  geom_line() +
  facet_grid(type~mean, scales = "free_y")
```


***Bonus  exercise***

There  is no reason the hazard function needs to follow one of these functions, or any function associated with a 'named' probability distribution.

Make up your own hazard function and calculate  the cumulative hazard, survival and density functions.

```{r}
# Your code goes here
```

