---
title: "Workbook for Introduction to TTE modeling"
author: "Introduction to TTE modeling"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  # html_document:
  #   css: docs/src/styles/styles.css
  #   number_sections: true
  #   theme: united
  #   toc: true
  #   toc_float: true
params:
  include: TRUE
---


```{r,echo=FALSE,message=FALSE}
knitr::opts_chunk$set(comment='.',fig.align=TRUE,message=FALSE,warning=FALSE)
```


# Preliminaries for R examples

```{r, message=FALSE}
library(tidyverse)
library(stringr)
library(survival)
library(survminer)
library(texreg)
library(mgcv)
```


# Plotting hazard and survival functions over time for some common distributions


## Exponential

The hazard is constant as a funtion of time: $h(t) = \lambda$.

From first principles:
 * h(t) = lambda
 * H(t) = lambda * t
 * S(t) = exp(-lambda * t)
 * f(t) = h(t) * S(t) = lambda * exp(-lambda * t)
 
```{r}
plot_data <- crossing(lambda = c(0.5,1,2), times = seq(0,5,length=100)) %>% 
  mutate(hazard = lambda,
         cumulative_hazard = lambda * times,
         survival = exp(-cumulative_hazard),
         density = hazard * survival)

plot_data %>% 
  pivot_longer(cols=hazard:density, names_to = 'type', values_to='value') %>% 
  ggplot(aes(x=times, y=value, col=as.factor(lambda), group=lambda)) +
  geom_line() +
  facet_wrap(~type, scales = "free_y")
```


## Weibull

The Weibull distribution has two parameters: lambda and gamma

The log hazard is linar in the log of time: $h(t) = \gamma \lambda t^{\gamma-1} \iff \log h(t) = \log\gamma + \log \lambda + (\gamma - 1) \log t$


From first principles:
 * h(t) = lambda * gamma * t^(gamma-1)
 * H(t) = lambda * t^gamma
 * S(t) = exp(-lambda * t^gamma)
 * f(t) = h(t) * S(t) = lambda * gamma * t^(gamma-1) * exp(-lambda * t^gamma)

```{r}
plot_data <- crossing(lambda = c(0.5,2), gamma = c(0.5, 1, 3), times = seq(0,5,length=100)) %>% 
  mutate(hazard = lambda * gamma * times^(gamma-1),
         cumulative_hazard = lambda * times^gamma,
         survival = exp(-cumulative_hazard),
         density = hazard * survival
)

plot_data %>% 
  pivot_longer(cols=hazard:density, names_to = 'type', values_to='value') %>% 
  mutate(combination = paste(lambda,gamma,sep='-')) %>% 
  ggplot(aes(x=times, y=value, col=as.factor(gamma), group=combination)) +
  geom_line() +
  facet_grid(type~lambda, scales = "free_y")
```
## Log-normal

The hazard isn't easily expressed in closed form.  How can you derive it from first principles?

Hint: the log normal pdf and survival function are obtained using the `dlnorm` and `plnorm` functions, respectively.

Plot the log-normal hazard function for combinations of mean (1, 2) and starndard deviation (0.3,1) modifying the code below.


```{r}
plot_data <- crossing(mean = c(1,2), sd = c(0.3, 1), times = seq(0,5,length=100)) %>% 
  mutate(survival = ,
         density = ,
         cumulative_hazard = ,
         hazard =
)

plot_data %>% 
  pivot_longer(cols=survival:hazard, names_to = 'type', values_to='value') %>% 
  mutate(combination = paste(mean,sd,sep='-')) %>% 
  ggplot(aes(x=times, y=value, col=as.factor(sd), group=combination)) +
  geom_line() +
  facet_grid(type~mean, scales = "free_y")
```



## Toy data

```{r, "TRY toydata", purl=TRUE, results = "hide"}
load('../data/aedat.RDS')
```

* This data set is comprised of a two week study (protocol A) and a six week study (protocol B)
* The data set includes patients of type `PT2`, a patient type that (let's say) has not been studied at higher dose levels.
* Suppose further that a dose level under consideration for a phase 2 study in `PT2` would have typical value exposures near 2.5 ug/mL.
* The objective of *this* anlaysis is to determine whether the AE rate will be "sufficiently low" in `PT2` at that exposure.



## EDA: data key

```{r, "TRY datakey", purl=TRUE, results = "hide"}

# Create a variable for time-to-serious AE's, matching "AE01"
aedat <-
  aedat %>% 
  mutate(AETOXGR = factor(aedat$AETOXGR, 0:3, labels=c("None","Mild","Moderate","Severe")),
         ae_any = AETOXGR != 'None') %>% 
  group_by(USUBJID) %>%
  # End of study for patients without a severe event
  mutate(TTE_SEVERE = case_when(
    STUDYID=="PROTA" ~ 2,
    STUDYID=="PROTB" ~ 6
  ),
  # Time of severe event for those that had one
  TTE_SEVERE = ifelse(AETOXGR=="Severe", TTE, TTE_SEVERE)
  )
```


## derived variables

Both for EDA and for model-checking, it's generally helpful to have quartiles of exposure:

```{r, "TRY varderive", purl=TRUE, results = "hide"}
dat_use <-
  aedat %>% arrange(USUBJID, TTE_SEVERE) %>% slice(1) %>%
  group_by(PBO) %>%
  mutate(Quartile = ifelse(PBO == "PBO", "PBO",
                           paste0("Q", ntile(CAVGSS, n = 4))))
```

## Estimating and plotting Kaplan-Meier estimates of survival distribution

The

```{r}
km_est = survfit(Surv(TTE,ae_any)~STUDYID, data = dat_use)
ggsurvplot(km_est, risk.table = TRUE)
```
Often it's helpful to add the number of subjects at risk to the bottom of the plot (a risk table).

```{r}
ggsurvplot(km_est, risk.table = TRUE)
```


Your turn:
 * Plot K-M estimates for quartiles of exposure (`Quartile`)


## Parametric survival model using Stan/brms

The censoring variable should contain the values 'left', 'none', 'right', and 'interval' (or equivalently -1, 0, 1, and 2) to indicate that the corresponding observation is left censored, not censored, right censored, or interval censored. For interval censored data, a second variable (let's call it y2) has to be passed to cens. In this case, the formula has the structure y | cens(censored, y2) ~ predictors. 


```{r, "TRY firststan", purl=TRUE, results = "hide"}
mod01_stan <- brm(formula = TTE | cens(1-ae_any) ~ Quartile,  
                  data=dat_use, 
                  family = weibull(),
                  warmup = 500, 
                  iter = 2000, 
                  chains = 4, 
                  inits = "0", 
                  cores = 4,
                  seed = 123)
```

Assess convergence (traceplots, rhat)

```{r}
rhat(mod01_stan)
```

```{r}
plot
```


Look at posterior distributions

Calculate model fit summary

```{r}
waic(mod01_stan)
```

Posterior predictive distributions

```{r}
# Generate posterior predictive samples (100 posterior samples)
post_preds <- dat_use %>% 
  add_predicted_draws(mod01_stan, n=100) %>%
  mutate(
    # TODO: Use a more general plug-in censoring function that is a function of data and .prediction
    # Censor using same process as in the source data
    .simdv = if_else(.prediction < STUDYDUR, 1, 0),
    .simtte = if_else(.prediction >= STUDYDUR, STUDYDUR,.prediction)) %>% 
  ungroup()

# Summarise posterior predictions at a grid of times
times = seq(0,6,length=100)

post_preds <- post_preds %>% 
  nest(sim_data = -.draw) %>% 
  bind_rows(dat_use %>% mutate(.draw=0, .simtte=TTE, .simdv=ae_any) %>% nest(sim_data = -.draw)) %>% 
  mutate(
    # Kaplan-Meier fits to each simulated dataset
    km = map(sim_data, ~survfit(Surv(.simtte, .simdv)~Quartile, data = .x)),
    # Extract predictions at a grid of times
    preds = map2(km, sim_data, function(.x,.y){
      .simfit = summary(.x, times=times)
      tibble(strata = .simfit$strata,
             surv = .simfit$surv,
             time = .simfit$time)
    })
  )

# Obtain median and quantiles for simulated data
post_pred_summary <- post_preds %>% 
  filter(.draw > 0) %>% 
  select(.draw, preds) %>% 
  unnest(cols=preds) %>% 
  group_by(strata, time) %>% 
  summarise(lpl = quantile(surv, probs = 0.05),
            upl = quantile(surv, probs = 0.95),
            surv = median(surv),
            n=n()) %>% 
  mutate(type = 'Simulated')


obs_summary <- post_preds %>% 
  filter(.draw == 0) %>% 
  select(preds) %>% 
  unnest(cols=preds) %>% 
  mutate(type='Observed')
  
  
  
bind_rows(obs_summary,post_pred_summary) %>% 
  ggplot(aes(x = time)) +
  geom_step(aes(y = surv, color=type)) +
  geom_ribbon(aes(ymin=lpl, ymax=upl, fill=type), alpha=0.25) +
  facet_wrap(~strata) +
  ylim(0,1)
  
```


```{r}
post_preds01 <- posterior_predict(mod01_stan)
```


Exercise:
* Re-fit the model using a probit link
* Re-fit the model with STUDYID as an additional covariate

```{r}

# * Re-fit the model using a probit link
brm(formula = AE01 ~ CAVGSS + BWT + PTTYPE + SEXTXT ,
    data=dat_mod,
    family = bernoulli(link = "probit"),
    warmup = 500,
    iter = 2000,
    chains = 4,
    inits = "0",
    cores = 4,
    seed = 123)

# * Re-fit the model with STUDYID as an additional covariate
# brm(formula = AE01 ~ CAVGSS + BWT + PTTYPE + SEXTXT + STUDYID,  
#     data=dat_mod, 
#     family = bernoulli(link = "logit"),
#     warmup = 500, 
#     iter = 2000, 
#     chains = 4, 
#     inits = "0", 
#     cores = 4,
#     seed = 123)

```

## reading R help

R help pages can be overwhelming ... but you have to learn to read them:  

```{r, "TRY extractors", purl=TRUE, results = "hide"}
?brm # See especially the "See Also" section
?summary.brmsfit # See expecially the "Value" section
?fixef
?predict
?predict.brmsfit
```

Exercise:

* Print out a table of estimated coefficients and their standard errors.
* Print out 95% confidence intervals for the model parameters.
* Print out 90% confidence intervals for the model parameters.

```{r}
fixef(mod01_stan)
fixef(mod01_stan,probs = c(0.05, 0.95))
```

## re-coding predictors

```{r, "TRY recoding", purl=TRUE, results = "hide"}
ref_bwt <- median(dat_use$BWT) # will want reference value later, so save as variable!
dat_mod <- dat_use %>% mutate(
  BWT_norm = BWT - ref_bwt,
  PTTYPE = relevel(PTTYPE, ref = c("PT1"))
  )
```

Exercise: re-fit the model so that the intercept corresponds to a `PT1` type patient with typical bodyweight.

```{r}

mod02_stan <- brm(formula = AE01 ~ CAVGSS + BWT_norm + PTTYPE + SEXTXT,  
                  data=dat_mod, 
                  family = bernoulli(link = "logit"),
                  warmup = 500, 
                  iter = 2000, 
                  chains = 4, 
                  inits = "0", 
                  cores = 4,
                  seed = 123)
```

Excercise:

* Transform the covariate coefficients to be more interpretable.
* How many fold higher are the odds of an AE for `PT1` versus `HV`?

```{r}
exp(fixef(mod02_stan)[,c("Estimate","Q2.5","Q97.5")])
# odds of an AE for HV versus PT1 are .21 lower
```

## residual plot

```{r, "TRY resid", purl=TRUE, results = "hide"}
dat_plus <- dat_mod
dat_plus$res <- residuals(mod02_stan)[,"Estimate"]
dat_plus$pred <- predict(mod02_stan)[,"Estimate"]

resplot <-
  ggplot(dat_plus, aes(x = CAVGSS, y = res)) +
  geom_point() +
  geom_smooth()

resplot  
resplot + facet_wrap(~ PTTYPE)
```


Extended Exercise:

* Re-fit the model to allow a different slope of exposure-response relationship for each level of `PTTYPE`.
* Re-generate the residual and pred v. conc plots and compare the results.
* Read `?aic` and compare the models with and without interaction in terms of AIC.

```{r}

mod03_stan <- brm(formula = AE01 ~ CAVGSS + BWT_norm + PTTYPE + SEXTXT + CAVGSS*PTTYPE,  
                  data=dat_mod, 
                  family = bernoulli(link = "logit"),
                  warmup = 500, 
                  iter = 2000, 
                  chains = 4, 
                  inits = "0", 
                  cores = 4,
                  seed = 123)

dat_plus <- dat_mod
dat_plus$res <- residuals(mod03_stan)[,"Estimate"]
dat_plus$pred <- predict(mod03_stan)[,"Estimate"]

resplot <-
  ggplot(dat_plus, aes(x = CAVGSS, y = res)) +
  geom_point() +
  geom_smooth()

resplot  
resplot + facet_wrap(~ PTTYPE)

waic1 <- waic(mod02_stan)
waic2 <- waic(mod03_stan)
loo_compare(waic1,waic2)

```

## summarizing interaction parameters

```{r, "TRY lincombo", purl=TRUE, results = "hide"}

posterior_samples(mod03_stan) %>% 
  transmute(lin_combo=`b_CAVGSS`+`b_CAVGSS:PTTYPEHV`) %>%
  with(list(est=mean(lin_combo),
            se=sd(lin_combo)))

```


# Slides03

## my first BRMS/STAN VPC

```{r, "TRY glmvpc", purl=TRUE, results = "hide"}
prop <- function(x) sum(x, na.rm = TRUE) / sum(!is.na(x))
obs_stat <-
  dat_mod %>%
  group_by(PTTYPE, Quartile) %>%
  summarise(pAE = prop(AE01))
ggplot(obs_stat) +
  geom_point(aes(x = Quartile, y = pAE)) +
  facet_wrap(~ PTTYPE)

nsim = 1000
sim_dat = posterior_predict(mod03_stan, nsamples = nsim)
sim_dat_long <- tibble(
  sim = rep(1:nrow(sim_dat), times = ncol(sim_dat)),
  id = rep(1:ncol(sim_dat), each = nrow(sim_dat)),
  value = c(sim_dat)
) %>% 
  left_join(mod03_stan$data %>% mutate(id=1:n()))

sim_dat_long %>% 
  group_by(PTTYPE,sim) %>% 
  summarise(phat=mean(value)) %>% 
  ggplot() + 
  geom_density(aes(x=phat, col=PTTYPE))

vpc.cat = function(cov)  {
  
  obs_stat <-
    dat_mod %>%
    group_by(across(all_of(c(cov,"Quartile")))) %>%
    summarise(pAE = prop(AE01))
  
  # Function to calculate predicted prob of event per PTTYPE and quartile
  stat_fn <- function(simy, dat_orig = dat_mod) {
    dati <- dat_orig
    dati$AE01 <- simy
    dati %>%
      group_by(across(all_of(c(cov,"Quartile")))) %>%
      summarise(pAE = prop(AE01))
  }
  
  sim_stat <- apply(sim_dat, 2, stat_fn) %>% bind_rows()
  
  sim_stat_q <-
    sim_stat %>%
    group_by(across(all_of(c(cov,"Quartile")))) %>%
    summarize(LB=quantile(pAE, prob=c(0.05)),
              MED=median(pAE),
              UB=quantile(pAE, prob=.95)
    )
  
  comb_stat <- full_join(obs_stat, sim_stat_q)
  
  ggplot(data = comb_stat) +
    geom_pointrange(aes(x = Quartile, y = MED, ymin = LB, ymax = UB, colour = "Predicted")) +
    geom_point(aes(x = Quartile, y = pAE, colour = "Observed"), size = 3, alpha = 0.5) +
    facet_wrap(~ get(cov)) + theme_bw()
}

vpc.cat("PTTYPE")
```


Exercise: Create a VPC for the proportion of patients with an AE within each gender.

```{r}
vpc.cat("SEXTXT")
```

## VPC against continuous predictors

```{r, "TRY continvpc", purl=TRUE, results = "hide"}


vpc.cont = function(cov) {
  
  obs_stat <-
    dat_mod %>%
    ungroup %>%
    mutate(pAE=predict(loess(AE01 ~ get(cov), data = dat_mod))) %>%
    group_by(across(all_of(cov))) %>%
    summarize(pAE=mean(pAE)) # Could use distinct, should be identical
  
  stat_fn <- function(simy, dat_orig = dat_mod) {
    dati <- dat_orig
    dati$AE01 <- simy
    dati$pAE <- predict(loess(AE01 ~ get(cov), data = dati))
    dati %>% group_by(across(all_of(cov))) %>% summarize(pAE=mean(pAE))
  }
  
  sim_stat <-
    data.frame(sim_dat) %>%
    select_all() %>%
    map(~stat_fn(.)) %>%
    bind_rows
  
  sim_stat_q <-
    sim_stat %>%
    group_by(across(all_of(cov))) %>%
    summarize(LB=quantile(pAE, prob=c(0.05)),
              MED=median(pAE),
              UB=quantile(pAE, prob=.95)
    )
  
  comb_stat <- full_join(obs_stat, sim_stat_q)
  
  ggplot(data = comb_stat) +
    geom_ribbon(aes(x = get(cov), ymin = LB, ymax = UB, colour = "Predicted"), fill = "lightgrey") +
    geom_line(aes(x = get(cov), y = MED, colour = "Predicted"), size = 3) +
    geom_line(aes(x = get(cov), y = pAE, colour = "Observed"), size = 3) +
    theme_bw()
  
}

vpc.cont("CAVGSS") + 
  labs(x="Concentraiton Average at Steady State",
       y="Probability of Adverse event",
       colour="")

```

Exercise: make a similar VPC plotted against bodyweight.

```{r}
vpc.cont("BWT") + 
  labs(x="Bodyweight",
       y="Probability of Adverse event",
       colour="")


```


Exercise: Our model includes an interacation term, make a VPC of CAVGSS by PTTYPE

```{r}

vpc.contcat = function(cont,
                       cat,
                       lev.cat) {
  
  dat_mod = dat_mod %>% ungroup()
  
  fun = function(lev,dat=dat_mod) {
    data.tmp = dat %>% ungroup() %>% filter(get(cat)==lev)
    data.tmp %>% mutate(pAE=predict(loess(AE01 ~ get(cont), data = data.tmp))) %>%
      group_by(across(all_of(c(cont,cat)))) %>% 
      summarize(pAE=mean(pAE)) 
  }
  
  obs_stat = map(lev.cat, fun) %>% bind_rows()
  
  stat_fn <- function(simy, dat_orig = dat_mod) {
    dati <- dat_orig
    dati$AE01 <- simy
    map(lev.cat,~fun(.x,dat=dati)) %>% 
      bind_rows()
  }
  
  sim_stat <-
    data.frame(sim_dat) %>%
    select_all() %>%
    map(~stat_fn(.)) %>%
    bind_rows
  
  sim_stat_q <-
    sim_stat %>%
    group_by(across(all_of(c(cont,cat)))) %>% 
    summarize(LB=quantile(pAE, prob=c(0.05)),
              MED=median(pAE),
              UB=quantile(pAE, prob=.95)
    )
  
  comb_stat <- full_join(obs_stat, sim_stat_q)
  
  comb_stat %>%  
    
    ggplot() +
    geom_ribbon(aes(x = get(cont), ymin = LB, ymax = UB, colour = "Predicted"), fill = "lightgrey") +
    geom_line(aes(x = get(cont), y = MED, group = get(cat), colour = "Predicted"), size = 3) +
    geom_line(aes(x = get(cont), y = pAE, group = get(cat), colour = "Observed"), size = 3) +
    facet_wrap(~ get(cat)) + theme_bw()
}

vpc.contcat("CAVGSS","PTTYPE",c("HV","PT1","PT2")) + 
  labs(x="Concentration Average at Steady State",
       y="Probability of Adverse event",
       colour="")

```

# Conditional predictions

```{r, "TRY forest", purl=TRUE, results = "hide"}
# We need some reference exposure levels:
ref_exp <- 1:3
# and some reference bodyweights:
ref_bwt_norm <- c(65, 75, 85) - ref_bwt
# For categorical covariates it is more straightforward:
ref_sex <- unique(dat_mod$SEXTXT)
ref_type <- unique(dat_mod$PTTYPE)
# Now let's make a grid of setting where we want predictions:
pred_frame <- expand.grid(BWT_norm = ref_bwt_norm,
                          CAVGSS = ref_exp,
                          SEXTXT = ref_sex,
                          PTTYPE = ref_type
                          )

qfun = function(x) {
  ret <- tibble::tibble(p = quantile(x, probs=c(.025,.5,.975),na.rm = TRUE))
  ret$q = c('LB','EST','UB')
  list(ret)
}

pred_frame0 = pred_frame
pred_frame = posterior_linpred(mod03_stan,newdata=pred_frame)

pred_frame = 
  plogis(pred_frame) %>% 
  data.frame() %>% 
  tibble() %>% pivot_longer(cols=everything(),
                            names_to="type",
                            values_to="value") %>% 
  group_by(type) %>% 
  summarise(across(c(value),list(qfun=~qfun(.x)))) %>% 
  unnest(value_qfun) %>% 
  pivot_wider(values_from=p,names_from=q) %>% 
  transmute(type=as.numeric(gsub("X","",type)),LB,EST,UB) %>% 
  arrange(type) %>% 
  cbind(pred_frame0) %>% 
  select(-type)


pred_frame$BWT <- round(pred_frame$BWT_norm + ref_bwt, 0)
pred_frame$Label = with(pred_frame, paste(PTTYPE, SEXTXT, BWT, CAVGSS))

pred_frame

ggplot(data = filter(pred_frame, PTTYPE != "HV")) +
  geom_pointrange(aes(x=Label, y=EST, ymin=LB, ymax=UB)) +
  coord_flip() + facet_wrap(~PTTYPE, scales="free_y") +
  xlab("Estimated AE Rate")

ggplot(data = filter(pred_frame, PTTYPE != "HV")) +
  geom_pointrange(aes(x=Label, y=EST, ymin=LB, ymax=UB)) +
  coord_flip() + facet_wrap(~SEXTXT, scales="free_y") +
  xlab("Estimated AE Rate")

```


Exercise: Re-fit the model without any covariate-expsoure interactions and re-make the "forest plot" of conditional predictions.

```{r}

# We need some reference exposure levels:
ref_exp <- 1:3
# and some reference bodyweights:
ref_bwt_norm <- c(65, 75, 85) - ref_bwt
# For categorical covariates it is more straightforward:
ref_sex <- unique(dat_mod$SEXTXT)
ref_type <- unique(dat_mod$PTTYPE)
# Now let's make a grid of setting where we want predictions:
pred_frame <- expand.grid(BWT_norm = ref_bwt_norm,
                          CAVGSS = ref_exp,
                          SEXTXT = ref_sex,
                          PTTYPE = ref_type
                          )

qfun = function(x) {
  ret <- tibble::tibble(p = quantile(x, probs=c(.025,.5,.975),na.rm = TRUE))
  ret$q = c('LB','EST','UB')
  list(ret)
}

pred_frame0 = pred_frame
#note using object mod02_stan instead of mod03_stan
pred_frame = posterior_linpred(mod02_stan,newdata=pred_frame)

pred_frame = 
  plogis(pred_frame) %>% 
  data.frame() %>% 
  tibble() %>% pivot_longer(cols=everything(),
                            names_to="type",
                            values_to="value") %>% 
  group_by(type) %>% 
  summarise(across(c(value),list(qfun=~qfun(.x)))) %>% 
  unnest(value_qfun) %>% 
  pivot_wider(values_from=p,names_from=q) %>% 
  transmute(type=as.numeric(gsub("X","",type)),LB,EST,UB) %>% 
  arrange(type) %>% 
  cbind(pred_frame0) %>% 
  select(-type)


pred_frame$BWT <- round(pred_frame$BWT_norm + ref_bwt, 0)
pred_frame$Label = with(pred_frame, paste(PTTYPE, SEXTXT, BWT, CAVGSS))

pred_frame

ggplot(data = filter(pred_frame, PTTYPE != "HV")) +
  geom_pointrange(aes(x=Label, y=EST, ymin=LB, ymax=UB)) +
  coord_flip() + facet_wrap(~PTTYPE, scales="free_y") +
  xlab("Estimated AE Rate")

ggplot(data = filter(pred_frame, PTTYPE != "HV")) +
  geom_pointrange(aes(x=Label, y=EST, ymin=LB, ymax=UB)) +
  coord_flip() + facet_wrap(~SEXTXT, scales="free_y") +
  xlab("Estimated AE Rate")

```

## Marginal predictions

When making marginal predictions, we wish to describe an average (marginal) effect over some population.

Let's compare marginal predictions of CAVGSS for PT1 and PT2 like those studied:

```{r, "TRY marginal-cavg-pttype", purl=TRUE}
dat_mod = dat_mod %>% ungroup()

pt1pats <- filter(dat_mod, PTTYPE=="PT1") %>%
  sample_n(1000,replace=TRUE) %>%
  select(CAVGSS,PTTYPE,BWT_norm,SEXTXT)

pt2pats <- filter(dat_mod, PTTYPE=="PT2") %>%
  sample_n(1000,replace=TRUE) %>%
  select(CAVGSS,PTTYPE,BWT_norm,SEXTXT)

pt1pats %>% ungroup %>% summary
pt2pats %>% ungroup %>% summary

CAVGSS_at <- seq(min(dat_mod$CAVGSS[dat_mod$CAVGSS>0]),
                 max(dat_mod$CAVGSS), length.out = 50)

pats <- bind_rows(pt1pats,pt2pats)%>% ungroup

pats$CAVGSS <- NULL

pred_eff <-
  map_df(seq_along(CAVGSS_at), function(CAVGSS_i){
    
    data.tmp =  pats %>% mutate(CAVGSS=CAVGSS_at[CAVGSS_i]) 
    
    posterior_linpred(mod03_stan, newdata=data.tmp) %>%
      plogis()%>% 
      data.frame() %>% 
      tibble() %>% pivot_longer(cols=everything(),
                                names_to="type",
                                values_to="value") %>% 
      group_by(type) %>% summarise(across(c(value),median)) %>% 
      transmute(type=as.numeric(gsub("X","",type)),value) %>% 
      arrange(type) %>% 
      cbind(data.tmp) %>%
      group_by(CAVGSS,PTTYPE) %>%
      summarize(pAE=median(value),
                ul=quantile(value,prob=.025),
                ll=quantile(value,prob=.975))
  })

pred_eff %>%
  ggplot(aes(x=CAVGSS, y=pAE, color=PTTYPE, fill=PTTYPE, ymax=ul, ymin=ll)) +
  geom_line() + theme_bw() +
  geom_ribbon(alpha=.25) +
  ggtitle("Marginal effect")
```

Exercise:
Suppose that exposure varies with bodyweight, plot the net effect across studied weights.

```{r}

dat_mod = dat_mod %>% mutate(BWT_Q=paste0("Q", ntile(BWT, n = 4))) 

pats = dat_mod %>% 
  group_by(BWT_Q) %>% 
  summarise(medBWT=median(BWT)) %>%
  right_join(dat_mod) %>% 
  group_by(BWT_Q) %>% 
  sample_n(1000,replace=TRUE) %>% 
  ungroup()

pats = pats %>% mutate(medBWT=factor(medBWT)) 

pats$CAVGSS <- NULL

pred_eff <-
  map_df(seq_along(CAVGSS_at), function(CAVGSS_i){
    
    data.tmp =  pats %>% mutate(CAVGSS=CAVGSS_at[CAVGSS_i]) 
    
    posterior_linpred(mod03_stan, newdata=data.tmp) %>%
      plogis()%>% 
      data.frame() %>% 
      tibble() %>% pivot_longer(cols=everything(),
                                names_to="type",
                                values_to="value") %>% 
      group_by(type) %>% summarise(across(c(value),median)) %>% 
      transmute(type=as.numeric(gsub("X","",type)),value) %>% 
      arrange(type) %>% 
      cbind(data.tmp) %>%
      group_by(CAVGSS,medBWT) %>%
      summarize(pAE=median(value),
                ul=quantile(value,prob=.025),
                ll=quantile(value,prob=.975))
  })

pred_eff %>%
  ggplot(aes(x=CAVGSS, y=pAE, color=medBWT, fill=medBWT, ymax=ul, ymin=ll)) +
  geom_line() + theme_bw() +
  geom_ribbon(alpha=.25) +
  ggtitle("Marginal effect")


```

## Clinical trial simulation

There is a planned Ph2 study in 100 PT2 patients.  
The patients will be dosed with a target CAVGSS of 2.5, and we expect values to fall
between 2.25 and 2.75 (say it varies normally around 2.5).  

* What rate of serious AE's should we expect in this trial?
* Compare with and without parameter uncertainty

Hint:
? MASS::mvrnorm
? model.matrix
? model.frame


```{r}

nsim = 100
# We need some reference exposure levels:
# range of CAVGSS if norm should approx 6 x sd 
ref_exp <- MASS::mvrnorm(nsim,mu=2.5,Sigma=(2.75-2.25)/6)
# and some reference bodyweights:
ref_bwt <- c(65, 75, 85)
# For categorical covariates it is more straightforward:
ref_sex <- unique(dat_mod$SEXTXT)
ref_type <- "PT1"
# Now let's make a grid of setting where we want predictions:
pred_frame <- expand.grid(BWT = ref_bwt,
                          CAVGSS = ref_exp,
                          SEXTXT = ref_sex,
                          PTTYPE = ref_type
                          )
pred_frame = pred_frame %>% 
  mutate(BWT_norm = BWT - ref_bwt,
         index=row_number())

pred_frame0 = pred_frame
pred_frame = posterior_linpred(mod03_stan,newdata=pred_frame) %>% 
  plogis() %>% 
  data.frame() %>%
  mutate(sim=row_number())

pred_frame = pred_frame %>% 
  tibble() %>% pivot_longer(cols=c(-sim),
                            names_to="index",
                            values_to="value") %>% 
  transmute(index=as.numeric(gsub("X","",index)),sim,value) %>% 
  left_join(pred_frame0)

pred_frame1 =  pred_frame %>% 
  group_by(index) %>% 
  summarise(across(c(value),median)) %>% 
  left_join(pred_frame0) %>% 
  group_by(CAVGSS,SEXTXT,BWT) %>% 
  summarize(pAE=median(value))
  
pred_frame1 %>%
  ggplot(aes(x=CAVGSS, y=pAE, )) +
  geom_line() + theme_bw() +
  facet_grid(SEXTXT~BWT) + 
  ggtitle("PAE without parameter uncertainty") + 
  ylim(c(0,1))

pred_frame2 =  pred_frame %>% 
  group_by(sim,CAVGSS,SEXTXT,BWT) %>% 
  summarise(across(c(value),median)) %>% 
  group_by(CAVGSS,SEXTXT,BWT) %>% 
      summarize(pAE=median(value),
                ul=quantile(value,prob=.025),
                ll=quantile(value,prob=.975))

pred_frame2 %>%
  ggplot(aes(x=CAVGSS, y=pAE, ), fill = "grey") +
  geom_line() + theme_bw() +
  geom_ribbon(aes(ymax=ul, ymin=ll),alpha=.25) +
  facet_grid(SEXTXT~BWT) + 
  ggtitle("PAE with parameter uncertainty") + 
  ylim(c(0,1))


```
























